---
title: "Endothelial.multiomics2"
output: html_document
date: "2025-07-11"
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo                  = FALSE
                      ,warning              = FALSE
                      ,message              = FALSE
                      ,error                = TRUE
                      ,dev                  = c('png','pdf')
                      ,pdf.options(encoding = "ISOLatin9.enc", useKerning=FALSE, useDingbats=FALSE))

```


A report by **Elyas.Mohammadi**

Here I try to prepare the scATACseq data from three sources. 1- Our in house generated data (Heart and brain) 2- Sten's data (Brain) 3- AmeenM data (Heart)

```{r libraries}

# Changing the Seurat version to 5
.libPaths("/cfs/klemming/projects/supr/naiss2023-6-180/elyasmo/Rloading/4.4.Spatial/")
# install.packages('Seurat')
# unloadNamespace("Seurat") 
library(ggplot2)
library(tibble)

library(Seurat)
library(stringr)
.libPaths("/cfs/klemming/projects/supr/naiss2023-6-180/elyasmo/Rloading/4.4")

# library(data.table)
library(harmony)

library(Signac)
library(ggrepel)
library(ggpubr)
library(patchwork)
# library(clusterProfiler)
# library(org.Hs.eg.db)
library(AnnotationHub)
# library(EnsDb.Hsapiens.v86)
library(ggthemes)
library(viridis)
library(dplyr)
library(GenomicRanges)
library(GenomeInfoDb)
library(tidyr)
library(RColorBrewer)
library(scales)

library(ggh4x)
axis <- ggh4x::guide_axis_truncated(
  trunc_lower = ggplot2::unit(0, "npc"),
  trunc_upper = ggplot2::unit(1.5, "cm"))

```


# Calculating the proportions

## Celltypes (L3)

```{r peaks}

ATAC.all.int <- readRDS(file="./ATAC.all.int.Splitted.Rds")
RNA.all.int <- readRDS(file="./RNA.int.Rds")

DefaultAssay(ATAC.all.int$Brain) <- "ACTIVITY"
DefaultAssay(ATAC.all.int$Heart) <- "ACTIVITY"
DefaultAssay(RNA.all.int) <- "RNA"

Idents(ATAC.all.int$Brain) <- "newL3_Cell_type.multi.pred"
Idents(ATAC.all.int$Heart) <- "newL3_Cell_type.multi.pred"
Idents(RNA.all.int) <- "newL3_Cell_type"



RNA.all.int <- subset(RNA.all.int, Age.PCW !="PCW22")
RNA.all.int <- SplitObject(RNA.all.int,split.by = "Organ")



## ---------- 1) Build tidy composition tables ----------
# ATAC: L3 column = newL3_Cell_type.multi.pred
brain_atac <- ATAC.all.int$Brain@meta.data %>%
  transmute(Organ = "Brain", Modality = "ATAC", L3 = newL3_Cell_type.multi.pred) %>%
  drop_na(L3) %>%
  count(Organ, Modality, L3, name = "n")

heart_atac <- ATAC.all.int$Heart@meta.data %>%
  transmute(Organ = "Heart", Modality = "ATAC", L3 = newL3_Cell_type.multi.pred) %>%
  drop_na(L3) %>%
  count(Organ, Modality, L3, name = "n")


# RNA: L3 column = newL3_Cell_type
brain_rna <- RNA.all.int$Brain@meta.data %>%
  transmute(Organ = "Brain", Modality = "RNA", L3 = newL3_Cell_type) %>%
  drop_na(L3) %>%
  count(Organ, Modality, L3, name = "n")

heart_rna <- RNA.all.int$Heart@meta.data %>%
  transmute(Organ = "Heart", Modality = "RNA", L3 = newL3_Cell_type) %>%
  drop_na(L3) %>%
  count(Organ, Modality, L3, name = "n")

# Combine and turn counts into proportions within Organ × Modality
comp_all <- bind_rows(brain_atac, heart_atac, brain_rna, heart_rna) %>%
  group_by(Organ, Modality) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

comp_all$L3 <- factor(comp_all$L3,levels = c("ECs_Cpillary" , "ECs_Arterial" , "Tip cells",  "Endocardial-like","ECs_Venous"  ,"ECs_Lymphatic" ,   "ECs_Mitotic", "Mural_Precursors" , "Mural_Fibroblasts" , "Mural_Pericytes" , "Mural_SMCs","Mural_Mitotic" ))


## ---------- 2) Color families (warm ECs, cool Mural) ----------
is_ec    <- function(x) grepl("^ECs_", x) | x %in% c("Tip cells", "Endocardial-like")
is_mural <- function(x) grepl("^Mural", x)


l3_levels <- sort(unique(comp_all$L3))
ec_l3     <- l3_levels[is_ec(l3_levels)]
mural_l3  <- l3_levels[is_mural(l3_levels)]
# warm ECs (skip the palest too, for balance)
ec_base    <- brewer.pal(9, "YlOrRd")
ec_cols    <- colorRampPalette(ec_base[2:9])(length(ec_l3))

# cool Mural (skip the near-white swatches)
mural_base <- brewer.pal(9, "PuBuGn")
mural_cols <- colorRampPalette(mural_base[3:9])(length(mural_l3))

fill_map <- c(setNames(ec_cols, ec_l3), setNames(mural_cols, mural_l3))
saveRDS(fill_map,"../Rsave/fill_map.Rds")

## ---------- 3) Plot helper (one organ at a time) ----------

fill_map <- readRDS("../Rsave/fill_map.Rds")

plot_organ <- function(df, organ_name) {
  d <- df %>%
    filter(Organ == organ_name) %>%
    mutate(
      L3 = factor(L3, levels = l3_levels),
      Modality = factor(Modality, levels = c("ATAC","RNA"))
    )
  ggplot(d, aes(x = Modality, y = prop, fill = L3)) +
    geom_col(width = 0.8) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    scale_fill_manual(values = fill_map, drop = TRUE, guide = guide_legend(ncol = 1)) +
    labs(
      title = organ_name,
      x = NULL, y = "Proportion (stacked to 100%)",
      fill = "L3 subtype"
    ) +
    theme_classic(base_size = 12) +
    theme(
      legend.position = "right",
      plot.title = element_text(face = "bold")
    )
}

## ---------- 4) Make the two plots ----------


p_brain <- plot_organ(comp_all, "Brain")+NoLegend()
p_heart <- plot_organ(comp_all, "Heart")+ylab("")

pdf("../Results/Manuscript/Figure1/Proportions_NOpcw22.pdf",width = 6,height = 8)
p_brain+p_heart
dev.off()

```


#Stacked.density

```{r Stacked.density,fig.height=8,fig.width=10}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(RColorBrewer)
library(patchwork)

num_age <- function(x) as.numeric(gsub("[^0-9.]", "", as.character(x)))

brain_atac <- ATAC.all.int$Brain@meta.data %>%
  transmute(Organ="Brain", Modality="ATAC", Age=num_age(Age.PCW),
            L3 = newL3_Cell_type.multi.pred) %>%
  drop_na(L3, Age)

heart_atac <- ATAC.all.int$Heart@meta.data %>%
  transmute(Organ="Heart", Modality="ATAC", Age=num_age(Age.PCW),
            L3 = newL3_Cell_type.multi.pred) %>%
  drop_na(L3, Age)

brain_rna <- RNA.all.int$Brain@meta.data %>%
  transmute(Organ="Brain", Modality="RNA", Age=num_age(Age.PCW),
            L3 = newL3_Cell_type) %>%
  drop_na(L3, Age)

heart_rna <- RNA.all.int$Heart@meta.data %>%
  transmute(Organ="Heart", Modality="RNA", Age=num_age(Age.PCW),
            L3 = newL3_Cell_type) %>%
  drop_na(L3, Age)

dat_all.list <- list(brain_rna, brain_atac, heart_rna, heart_atac)

pics1 <- lapply(dat_all.list[c(1,3)], function(df) {

  df$L3 <- factor(df$L3,
    levels = c("ECs_Cpillary","ECs_Arterial","Tip cells","Endocardial-like",
               "ECs_Venous","ECs_Lymphatic","ECs_Mitotic",
               "Mural_Precursors","Mural_Fibroblasts","Mural_Pericytes",
               "Mural_SMCs","Mural_Mitotic"))
  ggplot(df, aes(x = Age, fill = L3)) +
    geom_density(position = "fill", adjust = 1.5,
                 color = "grey90", linewidth = 0.2) +
    scale_y_continuous(labels = percent_format(accuracy = 1),
                       limits = c(0,1), expand = c(0,0)) +
    coord_cartesian(xlim = c(4,13)) +
    theme_classic(base_size = 12) +
    theme(
      legend.position = "none",
      plot.margin = margin(-6, 2, -6, 2),
      axis.title.x = element_blank(),
      axis.text.x  = element_blank(),
      axis.ticks.x = element_blank()
    )
})


pics2 <- lapply(dat_all.list[c(2,4)], function(df) {

  df$L3 <- factor(df$L3,
    levels = c("ECs_Cpillary","ECs_Arterial","Tip cells","Endocardial-like",
               "ECs_Venous","ECs_Lymphatic","ECs_Mitotic",
               "Mural_Precursors","Mural_Fibroblasts","Mural_Pericytes",
               "Mural_SMCs","Mural_Mitotic"))

  ggplot(df, aes(x = Age, fill = L3)) +
    geom_density(position = "fill", adjust = 1.5,
                 color = "grey90", linewidth = 0.2) +
    scale_y_continuous(labels = percent_format(accuracy = 1),
                       limits = c(0,1), expand = c(0,0)) +
    coord_cartesian(xlim = c(6,13)) +
    theme_classic(base_size = 12) +
    theme(
      legend.position = "none",
      plot.margin = margin(-6, 2, -6, 2),
      axis.title.x = element_blank(),
      axis.text.x  = element_blank(),
      axis.ticks.x = element_blank()
    )
})


fill_scale <- scale_fill_manual(values = fill_map,
                                limits = names(fill_map),
                                name = "L3 subtype")

p1 <- pics1[[1]] + fill_scale + ylab("Brain-RNA")   +
      theme(plot.margin = margin(0,2,-6,2))
p2 <- pics2[[1]] + fill_scale + ylab("Brain-ATAC")  +
      theme(plot.margin = margin(-6,2,-6,2))
p3 <- pics1[[2]] + fill_scale + ylab("Heart-RNA")   +
 theme(
        axis.text.x  = element_text(),
        axis.ticks.x = element_line(),
        plot.margin  = margin(-6,2,0,2),
        legend.position = "right"
      )
p4 <- pics2[[2]] + fill_scale + ylab("Heart-ATAC") +
      xlab("Age (PCW)") +
      theme(
        axis.text.x  = element_text(),
        axis.ticks.x = element_line(),
        plot.margin  = margin(-6,2,0,2),
        # legend.position = "right"
      )

p1 <- pics[[1]] + fill_scale + ylab("Brain-RNA") +
  theme(plot.margin = margin(6, 2, -6, 2))  # <- positive top margin

final1 <- (p1  / p3 ) +
  plot_layout(guides = "collect", heights = c(1,1)) 

final2 <- (p2  / p4 ) +
  plot_layout(guides = "collect", heights = c(1,1)) 



final <- wrap_plots(final1,final2,ncol = 1)


pdf("../Results/Manuscript/Figure1/Main2_NOpcw22.pdf",width = 9,height = 8)
final
dev.off()


pdf("../Results/Manuscript/Figure1/Main2.NOLEGEND_NOpcw22.pdf",width = 9,height = 8)
(pics[[1]]+ylab("Brain-RNA")+NoLegend())/(pics[[3]]+ylab("Brain-ATAC")+NoLegend())/(pics[[2]]+ylab("Heart-RNA")+NoLegend())/(pics[[4]]+ylab("Heart-ATAC")+xlab("Age (PCW)")+NoLegend())
dev.off()

```



# Dimplots for ATAC and RNA

```{r dims,fig.height=8,fig.width=10}
RNA.all.int <- readRDS(file="./RNA.int.Rds")
ATAC.all.int <- readRDS(file="./ATAC.all.int3.Rds")

RNA.all.int$newL3_Cell_type <- factor(RNA.all.int$newL3_Cell_type,levels = c("ECs_Cpillary" , "ECs_Arterial" , "Tip cells",  "Endocardial-like","ECs_Venous"  ,"ECs_Lymphatic" ,   "ECs_Mitotic", "Mural_Precursors" , "Mural_Fibroblasts" , "Mural_Pericytes" , "Mural_SMCs","Mural_Mitotic" ))

ATAC.all.int$newL3_Cell_type.multi.pred <- factor(ATAC.all.int$newL3_Cell_type.multi.pred,levels = c("ECs_Cpillary" , "ECs_Arterial" , "Tip cells",  "Endocardial-like","ECs_Venous"  ,"ECs_Lymphatic" ,   "ECs_Mitotic", "Mural_Precursors" , "Mural_Fibroblasts" , "Mural_Pericytes" , "Mural_SMCs","Mural_Mitotic" ))


p1 <- DimPlot(RNA.all.int,group.by = "newL3_Cell_type",cols = fill_map,reduction = "umap_harmony",split.by = "Organ",raster = F)+ggtitle("RNA")+xlab("UMAP 1")+ylab("UMAP 2")

p2 <- DimPlot(ATAC.all.int,group.by = "newL3_Cell_type.multi.pred",cols = fill_map,reduction = "int_umap",split.by = "Region",raster = F)+ggtitle("ATAC")+xlab("UMAP 1")+ylab("UMAP 2")


pdf("../Results/Manuscript/Figure1/Dimplot.SPL.pdf",width = 8,height = 10)
p1/p2
dev.off()


p1 <- DimPlot(RNA.all.int,group.by = "newL3_Cell_type",cols = fill_map,reduction = "umap_harmony",raster = F)+ggtitle("RNA")+xlab("UMAP 1")+ylab("UMAP 2")

p2 <- DimPlot(ATAC.all.int,group.by = "newL3_Cell_type.multi.pred",cols = fill_map,reduction = "int_umap",raster = F)+ggtitle("ATAC")+xlab("UMAP 1")+ylab("UMAP 2")

pdf("../Results/Manuscript/Figure1/Dimplot.pdf",width = 6,height = 8)
p1/p2
dev.off()


ATAC.all.int$Organ <- ATAC.all.int$Region
p1 <- DimPlot(RNA.all.int,group.by = "Organ",cols = c(Brain = "#3A86FF", Heart = "#2FBF71"),reduction = "umap_harmony",raster = F,split.by = "Age.PCW")+ggtitle("RNA")+xlab("UMAP 1")+ylab("UMAP 2")
p2 <- DimPlot(ATAC.all.int,group.by = "Organ",cols = c(Brain = "#3A86FF", Heart = "#2FBF71"),reduction = "int_umap",raster = F,split.by = "Age.PCW")+ggtitle("ATAC")+xlab("UMAP 1")+ylab("UMAP 2")


pdf("../Results/Manuscript/Final_refinments/ED1/Dimplot_timepoints.pdf",width = 17,height = 7)
p1/p2
dev.off()


ATAC.all.int$Source_Region <- gsub("Sten_Brain","MannensC.2023_Brain",ATAC.all.int$Source_Region)
ATAC.all.int$Source_Region <- gsub("AmeenM_Heart","AmeenM.2023_Heart",ATAC.all.int$Source_Region)
ATAC.all.int$Organ <-ATAC.all.int$Region
RNA.all.int$Dataset <- gsub("Mannens.Nat2024","MannensC.2023",RNA.all.int$Dataset)
RNA.all.int$Dataset <- gsub("Generated","Inhouse",RNA.all.int$Dataset)

RNA.all.int$Source_Region <- paste(RNA.all.int$Dataset,RNA.all.int$Organ,sep = "_")


p1 <- DimPlot(RNA.all.int,group.by = "Source_Region",reduction = "umap",raster = T)+labs(title = "RNA",subtitle = "Bfore integration")+xlab("UMAP 1")+ylab("UMAP 2")
p2 <- DimPlot(RNA.all.int,group.by = "Source_Region",reduction = "umap_harmony",raster = T)+labs(title = "RNA",subtitle = "After integration")+xlab("UMAP 1")+ylab("UMAP 2")
p3 <- p1|p2 

p1 <- DimPlot(ATAC.all.int,group.by = "Source_Region",reduction = "umap",raster = T)+labs(title = "ATAC",subtitle = "Bfore integration")+xlab("UMAP 1")+ylab("UMAP 2")
p2 <- DimPlot(ATAC.all.int,group.by = "Source_Region",reduction = "int_umap",raster = T)+labs(title = "ATAC",subtitle = "After integration")+xlab("UMAP 1")+ylab("UMAP 2")
p4 <- p1|p2 



pdf("../Results/Manuscript/Final_refinments/ED1/Dimplot_Before_After.int.pdf",width = 10,height = 7)
p3/p4
dev.off()


DefaultAssay(ATAC.all.int) <- "ACTIVITY"

RNA.all.int <- DietSeurat(RNA.all.int,assays = "RNA",dimreducs = F,graphs = F,scale.data = F)
ATAC.all.int <- DietSeurat(ATAC.all.int,assays = "ACTIVITY",dimreducs = F,graphs = F,scale.data = F)
RNA.all.int$Assay <- "RNA"
ATAC.all.int$Assay <- "ATAC"
ATAC.all.int <- RenameAssays(ATAC.all.int, ACTIVITY = "RNA")
DefaultAssay(ATAC.all.int) <- "RNA"

RNA.all.int$Batch2 <- RNA.all.int$Batch
ATAC.all.int$Batch2 <- ATAC.all.int$Source2
ATAC.all.int$newL3_Cell_type <- ATAC.all.int$newL3_Cell_type.multi.pred

merged <- merge(RNA.all.int,ATAC.all.int)

# 1) Split each object by Batch2
merged  <- SplitObject(merged,  split.by = "Batch2")


# 3) Pick shared features across ALL batches
features <- SelectIntegrationFeatures(merged, nfeatures = 2000)

# 4) Prepare each split for RPCA (scale+PCA on the same feature set)
for (i in seq_along(merged)) {
  merged[[i]] <- ScaleData(merged[[i]], features = features, verbose = FALSE)
  merged[[i]] <- RunPCA(  merged[[i]], features = features, npcs = 50, verbose = FALSE)
}

# 5) (Recommended) use RNA batches as references for stability
ref_idx <- which(names(merged) %in% unique(RNA.all.int$Batch2))

rm(RNA.all.int,ATAC.all.int)
gc()

anchors <- FindIntegrationAnchors(
  object.list     = merged,
  anchor.features = features,
  reduction       = "cca",
  # reference       = ref_idx,   # anchor to RNA splits
  dims            = 1:50
)

# 6) Integrate across batches + modalities
integrated <- IntegrateData(
  anchorset            = anchors,
  normalization.method = "LogNormalize",
  dims                 = 1:50
)

# ----------------------------
# 4) Build a shared embedding
# ----------------------------
DefaultAssay(integrated) <- "integrated"
integrated <- ScaleData(integrated, verbose = FALSE)
integrated <- RunPCA(integrated, npcs = 50, verbose = FALSE)

integrated <- RunUMAP(integrated, dims = 1320)

saveRDS(integrated,"../Rsave/integrated.ATAC.RNA2.Rds")

fill_map <- readRDS("../Rsave/fill_map.Rds")

DimPlot(integrated,group.by = "newL3_Cell_type",cols = fill_map,reduction = "umap",split.by = "Assay",raster = F)+ggtitle("Assay")+xlab("UMAP 1")+ylab("UMAP 2")
DimPlot(integrated,group.by = "Assay",raster = F)+ggtitle("RNA")+xlab("UMAP 1")+ylab("UMAP 2")



```


# Trying to find DEGs celltypes in RNA and ATAC (ACTIVITY assay) data

```{r load.deg,fig.width=5,fig.height=5}
ATAC.all.int <- readRDS(file="./ATAC.all.int.Splitted.Rds")
# ATAC.all.int <- readRDS(file="./ATAC.all.int.Splitted2.Rds")

RNA.brain.deg <- FindAllMarkers(RNA.all.int$Brain,logfc.threshold = 0.5,test.use = "MAST",only.pos = T,latent.vars = "Dataset")
RNA.heart.deg <- FindAllMarkers(RNA.all.int$Heart,logfc.threshold = 0.5,test.use = "MAST",only.pos = T)
ATAC.brain.deg <- FindAllMarkers(ATAC.all.int$Brain,logfc.threshold = 0.5,test.use = "MAST",only.pos = T,latent.vars = "Source2")
ATAC.heart.deg <- FindAllMarkers(ATAC.all.int$Heart,logfc.threshold = 0.5,test.use = "MAST",only.pos = T,latent.vars = "Source2")

save(ATAC.brain.deg,ATAC.heart.deg,RNA.brain.deg,RNA.heart.deg,file="./degs.RNA.ATAC.organ.L3.Rdata")
# save(ATAC.brain.deg,ATAC.heart.deg,RNA.brain.deg,RNA.heart.deg,file="./degs.RNA.ATAC.organ.L3_2.Rdata")

```


## Plotting the degs

```{r deg,fig.width=35,fig.height=12}

ATAC.all.int <- readRDS(file="./ATAC.all.int.Splitted.Rds")
RNA.all.int <- readRDS(file="./RNA.int.Rds")

DefaultAssay(ATAC.all.int$Brain) <- "ACTIVITY"
DefaultAssay(ATAC.all.int$Heart) <- "ACTIVITY"
DefaultAssay(RNA.all.int) <- "RNA"

Idents(ATAC.all.int$Brain) <- "newL3_Cell_type.multi.pred"
Idents(ATAC.all.int$Heart) <- "newL3_Cell_type.multi.pred"
Idents(RNA.all.int) <- "newL3_Cell_type"

RNA.all.int <- SplitObject(RNA.all.int,split.by = "Organ")


load(file="./degs.RNA.ATAC.organ.L3.Rdata")
RNA.all.int$Brain$newL3_Cell_type <- factor(RNA.all.int$Brain$newL3_Cell_type,levels = unique(RNA.all.int$Brain$newL3_Cell_type))
RNA.brain.deg$cluster <- factor(RNA.brain.deg$cluster,levels = levels(RNA.all.int$Brain$newL3_Cell_type))
RNA.brain.deg <- RNA.brain.deg[order(RNA.brain.deg$cluster),] 

RNA.all.int$Heart$newL3_Cell_type <- factor(RNA.all.int$Heart$newL3_Cell_type,levels = unique(RNA.all.int$Heart$newL3_Cell_type))
RNA.heart.deg$cluster <- factor(RNA.heart.deg$cluster,levels = levels(RNA.all.int$Heart$newL3_Cell_type))
RNA.heart.deg <- RNA.heart.deg[order(RNA.heart.deg$cluster),] 


ATAC.all.int$Brain$newL3_Cell_type.multi.pred <- factor(ATAC.all.int$Brain$newL3_Cell_type.multi.pred,levels = levels(RNA.all.int$Brain$newL3_Cell_type))
ATAC.brain.deg$cluster <- factor(ATAC.brain.deg$cluster,levels = levels(ATAC.all.int$Brain$newL3_Cell_type.multi.pred))
ATAC.brain.deg <- ATAC.brain.deg[order(ATAC.brain.deg$cluster),] 

ATAC.all.int$Heart$newL3_Cell_type.multi.pred <- factor(ATAC.all.int$Heart$newL3_Cell_type.multi.pred,levels = levels(RNA.all.int$Heart$newL3_Cell_type))
ATAC.heart.deg$cluster <- factor(ATAC.heart.deg$cluster,levels = levels(ATAC.all.int$Heart$newL3_Cell_type.multi.pred))
ATAC.heart.deg <- ATAC.heart.deg[order(ATAC.heart.deg$cluster),] 


ATAC.brain.deg <- subset(ATAC.brain.deg,avg_log2FC>0.7 & p_val_adj<0.01)
RNA.brain.deg <- subset(RNA.brain.deg,avg_log2FC>1 & p_val_adj<0.01)

ATAC.heart.deg <- subset(ATAC.heart.deg,avg_log2FC>0.7 & p_val_adj<0.01)
RNA.heart.deg <- subset(RNA.heart.deg,avg_log2FC>1 & p_val_adj<0.01)


tops.atac.brain <- c()
for (clusters in unique(ATAC.brain.deg$cluster)) {
  x <- subset(ATAC.brain.deg, cluster == clusters & p_val_adj<0.05)
  x <- x[order(-x$avg_log2FC, -x$p_val_adj),][1:5,]
  tops.atac.brain <- rbind(tops.atac.brain,x)
}

tops.atac.heart <- c()
for (clusters in unique(ATAC.heart.deg$cluster)) {
  x <- subset(ATAC.heart.deg, cluster == clusters & p_val_adj<0.05)
  x <- x[order(-x$avg_log2FC, -x$p_val_adj),][1:5,]
  tops.atac.heart <- rbind(tops.atac.heart,x)
}


tops.rna.brain <- c()
for (clusters in unique(RNA.brain.deg$cluster)) {
  x <- subset(RNA.brain.deg, cluster == clusters & p_val_adj<0.05)
  x <- x[order(-x$avg_log2FC, -x$p_val_adj),][1:5,]
  tops.rna.brain <- rbind(tops.rna.brain,x)
}

tops.rna.heart <- c()
for (clusters in unique(RNA.heart.deg$cluster)) {
  x <- subset(RNA.heart.deg, cluster == clusters & p_val_adj<0.05)
  x <- x[order(-x$avg_log2FC, -x$p_val_adj),][1:5,]
  tops.rna.heart <- rbind(tops.rna.heart,x)
}

##########################################################################################
# Removing cells with low number
ATAC.all.int$Brain <- subset(ATAC.all.int$Brain,newL3_Cell_type.multi.pred != "ECs_Mitotic")
ATAC.all.int$Heart <- subset(ATAC.all.int$Heart,newL3_Cell_type.multi.pred != "Mural_SMCs")
##########################################################################################


atac.brain <- DotPlot(ATAC.all.int$Brain,group.by = "newL3_Cell_type.multi.pred",features = unique(tops.atac.brain$gene),assay = "ACTIVITY",scale.by = "size")+scale_color_gradient2(low = "grey60",mid = "white",high = "purple3")+theme_bw()+theme(axis.text = element_text(face = "bold",size = 20),title = element_text(face = "bold",size = 20))+rotate_x_text(90)+ggtitle("ATAC-Brain")+xlab("")+ylab("Gene activity")

atac.heart <- DotPlot(ATAC.all.int$Heart,group.by = "newL3_Cell_type.multi.pred",features = unique(tops.atac.heart$gene),assay = "ACTIVITY",scale.by = "size")+scale_color_gradient2(low = "grey60",mid = "white",high = "purple3")+theme_bw()+theme(axis.text = element_text(face = "bold",size = 20),title = element_text(face = "bold",size = 20))+rotate_x_text(90)+ggtitle("ATAC-Heart")+ylab("Gene activity")


rna.brain <- DotPlot(RNA.all.int$Brain,group.by = "newL3_Cell_type",features = unique(tops.rna.brain$gene),assay = "RNA",scale.by = "size")+scale_color_gradient2(low = "grey60",mid = "white",high = "red")+theme_bw()+theme(axis.text = element_text(face = "bold",size = 20),title = element_text(face = "bold",size = 20))+rotate_x_text(90)+ggtitle("RNA-Brain")+xlab("")+ylab("Gene expression")

rna.heart <- DotPlot(RNA.all.int$Heart,group.by = "newL3_Cell_type",features = unique(tops.rna.heart$gene),assay = "RNA",scale.by = "size")+scale_color_gradient2(low = "grey60",mid = "white",high = "red")+theme_bw()+theme(axis.text = element_text(face = "bold",size = 20),title = element_text(face = "bold",size = 20))+rotate_x_text(90)+ggtitle("RNA-Heart")+ylab("Gene expression")



pdf("../Results/Manuscript/Figure1/Dotplot.Brain.pdf",width = 22,height = 12)
rna.brain/atac.brain
dev.off()

pdf("../Results/Manuscript/Figure1/Dotplot.Heart.pdf",width = 22,height = 12)
rna.heart/atac.heart
dev.off()



RNA.all.int <- readRDS(file="./RNA.int.Rds")

tops.rna.brain <- c()
for (clusters in unique(RNA.brain.deg$cluster)) {
  x <- subset(RNA.brain.deg, cluster == clusters & p_val_adj<0.05)
  x <- x[order(-x$avg_log2FC, -x$p_val_adj),][1:70,]
  tops.rna.brain <- rbind(tops.rna.brain,x)
}

tops.rna.heart <- c()
for (clusters in unique(RNA.heart.deg$cluster)) {
  x <- subset(RNA.heart.deg, cluster == clusters & p_val_adj<0.05)
  x <- x[order(-x$avg_log2FC, -x$p_val_adj),][1:70,]
  tops.rna.heart <- rbind(tops.rna.heart,x)
}

table(subset(tops.rna.heart,gene %in%unique(intersect(tops.rna.brain$gene,tops.rna.heart$gene)))$cluster)


tops.rna.brain <- subset(tops.rna.brain,gene %in% intersect(tops.rna.brain$gene,tops.rna.heart$gene))

tops.rna.heart <- subset(tops.rna.heart,gene %in% intersect(tops.rna.brain$gene,tops.rna.heart$gene))

ISS <- readxl::read_excel("../Data/Gene list for ISS (175).xlsx")
ISS <- intersect(ISS$`Gene Symbol`,tops.rna.brain$gene)

tops.rna.brain2 <- c()
for (clusters in unique(tops.rna.brain$cluster)) {
  x <- subset(tops.rna.brain, cluster == clusters & p_val_adj<0.05)
  x <- x[order(-x$avg_log2FC, -x$p_val_adj),][1:7,]
  tops.rna.brain2 <- rbind(tops.rna.brain2,x)
}

tops.rna.heart2 <- c()
for (clusters in unique(tops.rna.heart$cluster)) {
  x <- subset(tops.rna.heart, cluster == clusters & p_val_adj<0.05)
  x <- x[order(-x$avg_log2FC, -x$p_val_adj),][1:7,]
  tops.rna.heart2 <- rbind(tops.rna.heart2,x)
}

table(subset(tops.rna.heart2,gene %in%unique(intersect(tops.rna.brain2$gene,tops.rna.heart2$gene)))$cluster)


RNA.all.int$newL3_Cell_type <- factor(RNA.all.int$newL3_Cell_type,levels = c(levels(tops.rna.brain2$cluster)))


DotPlot(RNA.all.int,group.by = "newL3_Cell_type",split.by = "Organ",features = unique(intersect(tops.rna.brain2$gene,tops.rna.heart2$gene)),assay = "RNA",scale.by = "size",cols = c("Brain"="red4","Heart"="red4"))+theme_bw()+theme(axis.text = element_text(face = "bold",size = 20),title = element_text(face = "bold",size = 20))+rotate_x_text(90)+ggtitle("ATAC-Brain")+xlab("")+ylab("Gene activity")




fill_map <- readRDS("../Rsave/fill_map.Rds")

tops.rna.brain2$cluster <- factor(tops.rna.brain2$cluster,levels = names(fill_map))
tops.rna.heart2$cluster <- factor(tops.rna.heart2$cluster,levels = names(fill_map))

tops.rna.brain2 <- tops.rna.brain2[order(tops.rna.brain2$cluster),]
tops.rna.heart2 <- tops.rna.heart2[order(tops.rna.heart2$cluster),]
tops.rna.brain2 <- tops.rna.brain2[complete.cases(tops.rna.brain2),]
tops.rna.heart2 <- tops.rna.heart2[complete.cases(tops.rna.heart2),]

genes_use <- na.omit(c(ISS,unique(intersect(tops.rna.brain2$gene, tops.rna.heart2$gene))))

RNA.all.int$Organ_Celltype <- paste(RNA.all.int$Organ, RNA.all.int$newL3_Cell_type, sep="__")
avg <- AverageExpression(
  RNA.all.int,
  features = genes_use,
  group.by = "Organ_Celltype",
  assays = "RNA",
)$RNA

row_info <- data.frame(
  Organ = sub("--.*", "", colnames(avg)),
  CellType = sub(".*--", "", colnames(avg)),
  row.names = colnames(avg)
)

names(fill_map) <- gsub("-","_",names(fill_map))

ct_order <- names(fill_map)
row_info$CellType <- gsub("-","_",row_info$CellType)

row_info$CellType <- factor(row_info$CellType, levels = ct_order)
row_info <- row_info[order(row_info$CellType),]

avg <- avg[,rownames(row_info)]


ann_colors <- list(
  Organ = c(Brain = "#3A86FF", Heart = "#2FBF71"),
  CellType = fill_map[ct_order]
)

my_colors <- colorRampPalette(c("grey20","grey20","grey20", "white","red1","red3","red3"))(101)

pdf("../Results/Manuscript/Figure1/Shared.genes.full.pdf",width = 10,height = 15)

pheatmap::pheatmap(
  avg,
  color = my_colors,
  breaks = seq(-2,2,length.out=102),
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = row_info,
  annotation_colors = ann_colors,
  scale = "row",
  show_colnames = FALSE,
  fontsize_row = 7
)
dev.off()
  

selected <- c("AL162718.1","ARHGEF28"  ,"ACP7","LINC00841","CDH13", "P2RY8","LINC01896", "SALL3","ADGRG6","SELP", "CSF3","SELE"
,"ICAM1","STC1","TFF3", "LYVE1","RRM2","CDK1","HIST1H1B","CDC20","PLK1","TOP2A","MKI67","BRINP3","OGN","ALK","CCDC102B","KCNJ8","ABCC9","AC079822.1","OR51E1","MYH11",
"MYOCD","ACTG2","TAGLN","ACTA2","AC005394.1")

pdf("../Results/Manuscript/Figure1/Shared.genes.subset.pdf",width = 10,height = 12)

pheatmap::pheatmap(
  avg[selected,],
  color = my_colors,
  breaks = seq(-2,2,length.out=102),
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = row_info,
  annotation_colors = ann_colors,
  scale = "row",
  show_colnames = FALSE,
  fontsize_row = 7
)
dev.off()

```




## Selecting the DEGs that are shared between brain and heart celltypes


```{r shared.signatuers,fig.width=25,fig.height=12}

ISS <- readxl::read_excel("../Data/Gene list for ISS (175).xlsx")

ATAC.brain.deg.ISS <- subset(ATAC.brain.deg,gene%in% ISS$`Gene Symbol`)
RNA.brain.deg.ISS <- subset(RNA.brain.deg,gene%in% ISS$`Gene Symbol`)
ATAC.heart.deg.ISS <- subset(ATAC.heart.deg,gene%in% ISS$`Gene Symbol`)
RNA.heart.deg.ISS <- subset(RNA.heart.deg,gene%in% ISS$`Gene Symbol`)


atac.brain <- DotPlot(ATAC.all.int$Brain,group.by = "newL3_Cell_type.multi.pred",features = unique(ATAC.brain.deg.ISS$gene),assay = "ACTIVITY",scale.by = "size")+scale_color_gradient2(low = "grey60",mid = "white",high = "purple3")+theme_bw()+theme(axis.text = element_text(face = "bold",size = 20),title = element_text(face = "bold",size = 20))+rotate_x_text(90)+ggtitle("ATAC-Brain(ISS)")+xlab("")+ylab("Gene activity")

atac.heart <- DotPlot(ATAC.all.int$Heart,group.by = "newL3_Cell_type.multi.pred",features = unique(ATAC.heart.deg.ISS$gene),assay = "ACTIVITY",scale.by = "size")+scale_color_gradient2(low = "grey60",mid = "white",high = "purple3")+theme_bw()+theme(axis.text = element_text(face = "bold",size = 20),title = element_text(face = "bold",size = 20))+rotate_x_text(90)+ggtitle("ATAC-Heart(ISS)")+ylab("Gene activity")


rna.brain <- DotPlot(RNA.all.int$Brain,group.by = "newL3_Cell_type",features = unique(RNA.brain.deg.ISS$gene),assay = "RNA",scale.by = "size")+scale_color_gradient2(low = "grey60",mid = "white",high = "red")+theme_bw()+theme(axis.text = element_text(face = "bold",size = 20),title = element_text(face = "bold",size = 20))+rotate_x_text(90)+ggtitle("RNA-Brain(ISS)")+xlab("")+ylab("Gene expression")

rna.heart <- DotPlot(RNA.all.int$Heart,group.by = "newL3_Cell_type",features = unique(RNA.heart.deg.ISS$gene),assay = "RNA",scale.by = "size")+scale_color_gradient2(low = "grey60",mid = "white",high = "red")+theme_bw()+theme(axis.text = element_text(face = "bold",size = 20),title = element_text(face = "bold",size = 20))+rotate_x_text(90)+ggtitle("RNA-Heart(ISS)")+ylab("Gene expression")



pdf("../Results/Manuscript/Figure1/Dotplot.Brain.shared(ISS).pdf",width = 40,height = 12)
rna.brain/atac.brain
dev.off()

pdf("../Results/Manuscript/Figure1/Dotplot.Heart.shared(ISS).pdf",width = 40,height = 12)
rna.heart/atac.heart
dev.off()

pdf("../Results/Manuscript/Figure1/Dotplot.Brain.Heart.shared(ISS).pdf",width = 50,height = 12)
rna.brain/rna.heart
dev.off()



tops.atac.brain <- c()
for (clusters in unique(ATAC.brain.deg$cluster)) {
  x <- subset(ATAC.brain.deg, cluster == clusters & p_val_adj<0.05)
  x <- x[order(-x$avg_log2FC, -x$p_val_adj),][1:100,]
  tops.atac.brain <- rbind(tops.atac.brain,x)
}

tops.atac.heart <- c()
for (clusters in unique(ATAC.heart.deg$cluster)) {
  x <- subset(ATAC.heart.deg, cluster == clusters & p_val_adj<0.05)
  x <- x[order(-x$avg_log2FC, -x$p_val_adj),][1:100,]
  tops.atac.heart <- rbind(tops.atac.heart,x)
}


tops.rna.brain <- c()
for (clusters in unique(RNA.brain.deg$cluster)) {
  x <- subset(RNA.brain.deg, cluster == clusters & p_val_adj<0.05)
  x <- x[order(-x$avg_log2FC, -x$p_val_adj),][1:100,]
  tops.rna.brain <- rbind(tops.rna.brain,x)
}

tops.rna.heart <- c()
for (clusters in unique(RNA.heart.deg$cluster)) {
  x <- subset(RNA.heart.deg, cluster == clusters & p_val_adj<0.05)
  x <- x[order(-x$avg_log2FC, -x$p_val_adj),][1:100,]
  tops.rna.heart <- rbind(tops.rna.heart,x)
}

# 
# Brain.add <- intersect(intersect(tops.rna.brain$gene,ISS$`Gene Symbol`),intersect(tops.atac.brain$gene,ISS$`Gene Symbol`))
# Heart.add <- intersect(intersect(tops.rna.heart$gene,ISS$`Gene Symbol`),intersect(tops.atac.heart$gene,ISS$`Gene Symbol`))

  
# BRAIN <- c("DPEP1","DEPP1","EGFL8","IFI27","PLVAP","MMP1","MMP10","ESM1","NLGN4Y","HIST1H3B","MSX2","FZD2","COL11A1","PAX3","TOX3","TSLP","NUPR1","OGN","CRABP1","CCBE1",
# "SYP", #Endocardial-like?    
# "ACKR1","IL33","HIGD1B","FHL5","SULT1E1","CRYAB","TAGLN","WFDC1","ACAN","MYL9","ACTA2")
# 
# HEART <- c("KLHL6","CA4","EGFL8","TSPAN2","GPIHBP1","PRND","ACKR1","MEDAG","CD69","MADCAM1","OIT3","LXN","SLC16A13","TNFSF9","KCNJ8","COL5A3","TFF3","FLT4","TBX1","LYVE1","LAPTM5","PDPN","MRC1","STON2","STAB2","HIST1H1D","HIST1H2AL","HIST1H2AJ","TYMS","OMD","TNFAIP6","LUM","PRRX1","SMOC1","MYOZ2","CYSLTR2","AVPR1A","FOXF1","PTTG1","EGR4","PSRC1")


BRAIN <- c("DPEP1","ABCB1","DEPP1","EGFL8","IFI27","GJA4","PLVAP","MMP1","MMP10","ESM1","NLGN4Y","HPGD", "TFPI2", "CLEC2B", "CETP", "TMEM233","HIST1H3B","MSX2","FZD2","COL11A1","NUSAP1","CDC20","PAX3","TOX3","TSLP","RFX4","NUPR1","OGN","CRABP1","CCBE1",
"SYP", #Endocardial-like?    
"ACKR1","IL33","ADGRG6","HIGD1B","FHL5","SULT1E1","CRYAB","TAGLN","WFDC1","ACAN","MYL9","ACTA2","ACTG2","TAGLN","MYOCD","ACTA2","COL8A1")

HEART <- c("KLHL6","CA4","CLIC5","EGFL8","TSPAN2","GPIHBP1","PRND","ACKR1","MEDAG","CD69","MADCAM1","OIT3","LXN","SLC16A13","TNFSF9","ADM","IER5","TFF3","FLT4","TBX1","LYVE1","LAPTM5","PDPN","MRC1","STON2","STAB2","HIST1H1D","HIST1H2AL","HIST1H2AJ","TYMS","OMD","TNFAIP6","LUM","PRRX1","SMOC1","MYOZ2","CYSLTR2","AVPR1A","CCDC102B","KCNJ8","COL5A3","FOXF1","PTTG1","EGR4","PSRC1")

# 
# Brain.add= c("ABCB1" ,"GJA4", "HPGD", "TFPI2", "CLEC2B", "CETP", "TMEM233","NUSAP1","CDC20","RFX4","ADGRG6","ACTG2","TAGLN","MYOCD","ACTA2","COL8A1")
# Heart.add= c("CLIC5","ADM","IER5","HIST1H1B","TYMS", "HIST1H1D","CCDC102B","LYVE1","TFF3",)


# tops.atac.brain <- subset(ATAC.brain.deg, gene%in%unique(c(BRAIN,Brain.add)))
# tops.atac.heart <- subset(tops.atac.heart, gene%in%unique(c(HEART,Heart.add)))
# tops.rna.brain <- subset(RNA.brain.deg, gene%in%unique(c(BRAIN,Brain.add)))
# tops.rna.heart <- subset(tops.rna.heart, gene%in%unique(c(HEART,Heart.add)))
# 
# tops.atac.brain <- tops.atac.brain %>%
#   group_by(gene) %>%
#   slice_max(order_by = avg_log2FC,n=1,with_ties = FALSE) %>%
#   ungroup()
# 
# tops.atac.heart <- tops.atac.heart %>%
#   group_by(gene) %>%
#   slice_max(order_by = avg_log2FC,n=1,with_ties = FALSE) %>%
#   ungroup()
# 
# tops.rna.brain <- tops.rna.brain %>%
#   group_by(gene) %>%
#   slice_max(order_by = avg_log2FC,n=1,with_ties = FALSE) %>%
#   ungroup()
# 
# tops.rna.heart <- tops.rna.heart %>%
#   group_by(gene) %>%
#   slice_max(order_by = avg_log2FC,n=1,with_ties = FALSE) %>%
#   ungroup()
# 
# 
# tops.atac.brain <- tops.atac.brain[order(tops.atac.brain$cluster),] 
# tops.atac.heart <- tops.atac.heart[order(tops.atac.heart$cluster),] 
# tops.rna.brain <- tops.rna.brain[order(tops.rna.brain$cluster),] 
# tops.rna.heart <- tops.rna.heart[order(tops.rna.heart$cluster),] 


atac.brain <- DotPlot(ATAC.all.int$Brain,group.by = "newL3_Cell_type.multi.pred",features = unique(BRAIN),assay = "ACTIVITY",scale.by = "size")+scale_color_gradient2(low = "grey60",mid = "white",high = "purple3")+theme_bw()+theme(axis.text = element_text(face = "bold",size = 20),title = element_text(face = "bold",size = 20))+rotate_x_text(90)+ggtitle("ATAC-Brain")+xlab("")+ylab("Gene activity")

atac.heart <- DotPlot(ATAC.all.int$Heart,group.by = "newL3_Cell_type.multi.pred",features = unique(HEART),assay = "ACTIVITY",scale.by = "size")+scale_color_gradient2(low = "grey60",mid = "white",high = "purple3")+theme_bw()+theme(axis.text = element_text(face = "bold",size = 20),title = element_text(face = "bold",size = 20))+rotate_x_text(90)+ggtitle("ATAC-Heart")+ylab("Gene activity")


rna.brain <- DotPlot(RNA.all.int$Brain,group.by = "newL3_Cell_type",features = unique(BRAIN),assay = "RNA",scale.by = "size")+scale_color_gradient2(low = "grey60",mid = "white",high = "red")+theme_bw()+theme(axis.text = element_text(face = "bold",size = 20),title = element_text(face = "bold",size = 20))+rotate_x_text(90)+ggtitle("RNA-Brain")+xlab("")+ylab("Gene expression")

rna.heart <- DotPlot(RNA.all.int$Heart,group.by = "newL3_Cell_type",features = unique(HEART),assay = "RNA",scale.by = "size")+scale_color_gradient2(low = "grey60",mid = "white",high = "red")+theme_bw()+theme(axis.text = element_text(face = "bold",size = 20),title = element_text(face = "bold",size = 20))+rotate_x_text(90)+ggtitle("RNA-Heart")+ylab("Gene expression")



pdf("../Results/Manuscript/Figure1/Dotplot.Brain.shared.pdf",width = 20,height = 12)
rna.brain/atac.brain
dev.off()

pdf("../Results/Manuscript/Figure1/Dotplot.Heart.shared.pdf",width = 20,height = 12)
rna.heart/atac.heart
dev.off()


# table(tops.atac.brain[tops.atac.brain$gene%in%RNA.brain.deg$gene,]$cluster)
# table(RNA.brain.deg[RNA.brain.deg$gene %in%tops.atac.brain$gene,]$cluster)
# 
# 
# table(tops.atac.heart[tops.atac.heart$gene%in%RNA.heart.deg$gene,]$cluster)
# table(RNA.heart.deg[RNA.heart.deg$gene %in%tops.atac.heart$gene,]$cluster)


```


we can combine the four dotplots about in the one general heatmap

```{r heatmap.fig1,fig.width=25,fig.height=12}


avg.brain.rna <- as.data.frame(AverageExpression(RNA.all.int$Brain,features = unique(c(BRAIN)),assays = "RNA",group.by = "newL3_Cell_type"))
avg.brain.atac <- as.data.frame(AverageExpression(ATAC.all.int$Brain,assays = "ACTIVITY",features = unique(c(BRAIN)),group.by = "newL3_Cell_type.multi.pred"))
avg.heart.rna <- as.data.frame(AverageExpression(RNA.all.int$Heart,features = unique(c(HEART)),assays = "RNA",group.by = "newL3_Cell_type"))
avg.heart.atac <- as.data.frame(AverageExpression(ATAC.all.int$Heart,assays = "ACTIVITY",features = unique(c(HEART)),group.by = "newL3_Cell_type.multi.pred"))



## --- 1) Row-wise min–max scaler to [-1, 1] (per matrix) ---
scale_rows_to_pm1 <- function(M) {
  # M: genes x samples
  scale_row <- function(x) {
    rmin <- min(x, na.rm = TRUE); rmax <- max(x, na.rm = TRUE)
    if (!is.finite(rmin) || !is.finite(rmax) || rmax == rmin) {
      return(rep(0, length(x)))             # flat row -> all zeros
    }
    2 * ((x - rmin) / (rmax - rmin)) - 1    # map to [-1, 1]
  }
  t(apply(M, 1, scale_row))
}


avg.brain.rna <- scale_rows_to_pm1(avg.brain.rna)
avg.brain.atac <- scale_rows_to_pm1(avg.brain.atac)
avg.heart.rna <- scale_rows_to_pm1(avg.heart.rna)
avg.heart.atac <- scale_rows_to_pm1(avg.heart.atac)


matrix1 <- as.data.frame(cbind(avg.brain.rna,avg.brain.atac))
matrix2 <-as.data.frame(cbind(avg.heart.rna,avg.heart.atac))

matrix1$ACTIVITY.ECs.Lymphatic <- 0
matrix1$ACTIVITY.ECs.Mitotic <- 0
matrix2$ACTIVITY.Mural.SMCs <- 0

matrix2 <- matrix2[,match(colnames(matrix1),colnames(matrix2))]


df <- data.frame(Names=c(paste0("Brain_",colnames(matrix1)),
                         paste0("Heart_",colnames(matrix2))))


rownames(df) <- df$Names
df$Names <- gsub("Brain_RNA.","",df$Names)
df$Names <- gsub("Brain_ACTIVITY.","",df$Names)
df$Names <- gsub("Heart_RNA.","",df$Names)
df$Names <- gsub("Heart_ACTIVITY.","",df$Names)
df$Organ <- gsub("_.*","",rownames(df))
df$Method <- ifelse(grepl("RNA",rownames(df))=="TRUE","RNA","ATAC")

df$Organ <- factor(df$Organ,levels = c("Brain","Heart"))
df$Method <- factor(df$Method,levels = c("RNA","ATAC"))

cell.order <- subset(df,Organ=="Heart")$Name[1:12]
cell.order <- gsub("\\.", "_", cell.order)
cell.order <- gsub("Tip_cells", "Tip cells", cell.order)
cell.order <- gsub("Endocardial_like", "Endocardial-like", cell.order)

RNA.heart.deg$cluster <- factor(RNA.heart.deg$cluster,levels = cell.order)
RNA.heart.deg <- RNA.heart.deg[order(RNA.heart.deg$cluster),] 
RNA.heart.deg <- subset(RNA.heart.deg,gene %in% HEART)

ATAC.heart.deg$cluster <- factor(ATAC.heart.deg$cluster,levels = cell.order)
ATAC.heart.deg <- ATAC.heart.deg[order(ATAC.heart.deg$cluster),] 
ATAC.heart.deg <- subset(ATAC.heart.deg,gene %in% HEART)



### Here we could define the correct order of heart genes for the heatmap
HEART <- unique(ATAC.heart.deg$gene)

ATAC.all.int$Heart$newL3_Cell_type.multi.pred <- factor(ATAC.all.int$Heart$newL3_Cell_type.multi.pred,levels = levels(ATAC.all.int$Heart))
RNA.all.int$Heart$newL3_Cell_type <- factor(RNA.all.int$Heart$newL3_Cell_type ,levels = levels(RNA.all.int$Heart))

avg.brain.rna <- as.data.frame(AverageExpression(RNA.all.int$Brain,features = unique(c(BRAIN)),assays = "RNA",group.by = "newL3_Cell_type"))
avg.brain.atac <- as.data.frame(AverageExpression(ATAC.all.int$Brain,assays = "ACTIVITY",features = unique(c(BRAIN)),group.by = "newL3_Cell_type.multi.pred"))
avg.heart.rna <- as.data.frame(AverageExpression(RNA.all.int$Heart,features = unique(c(HEART)),assays = "RNA",group.by = "newL3_Cell_type"))
avg.heart.atac <- as.data.frame(AverageExpression(ATAC.all.int$Heart,assays = "ACTIVITY",features = unique(c(HEART)),group.by = "newL3_Cell_type.multi.pred"))

## --- 1) Row-wise min–max scaler to [-1, 1] (per matrix) ---
scale_rows_to_pm1 <- function(M) {
  # M: genes x samples
  scale_row <- function(x) {
    rmin <- min(x, na.rm = TRUE); rmax <- max(x, na.rm = TRUE)
    if (!is.finite(rmin) || !is.finite(rmax) || rmax == rmin) {
      return(rep(0, length(x)))             # flat row -> all zeros
    }
    2 * ((x - rmin) / (rmax - rmin)) - 1    # map to [-1, 1]
  }
  t(apply(M, 1, scale_row))
}


avg.brain.rna <- scale_rows_to_pm1(avg.brain.rna)
avg.brain.atac <- scale_rows_to_pm1(avg.brain.atac)
avg.heart.rna <- scale_rows_to_pm1(avg.heart.rna)
avg.heart.atac <- scale_rows_to_pm1(avg.heart.atac)


matrix1 <- as.data.frame(cbind(avg.brain.rna,avg.brain.atac))
matrix2 <-as.data.frame(cbind(avg.heart.rna,avg.heart.atac))

matrix1$ACTIVITY.ECs.Lymphatic <- 0
matrix1$ACTIVITY.ECs.Mitotic <- 0
matrix2$ACTIVITY.Mural.SMCs <- 0

matrix2 <- matrix2[,match(colnames(matrix1),colnames(matrix2))]

Matrix <- rbind(matrix1,matrix2)
Matrix <- Matrix %>%
  relocate(ACTIVITY.ECs.Mitotic, .before = ACTIVITY.Mural.Mitotic)

Matrix <- t(Matrix)
colnames(Matrix) <- gsub("EGFL81","EGFL8",colnames(Matrix))
colnames(Matrix) <- gsub("ACKR11","ACKR1",colnames(Matrix))

Matrix <- as.data.frame(Matrix) %>%
  relocate(all_of(c("LXN","OIT3")), .before = "CD69")

Matrix <- as.data.frame(Matrix) %>%
  relocate(all_of(c("LYVE1" , "FLT4","LAPTM5")), .before = "TBX1")

Matrix <- as.data.frame(Matrix) %>%
  relocate(all_of(c("ACTA2" )), .before = "LYVE1")

library(pheatmap)

## 1) Keep a copy of the original (possibly duplicated) gene labels
orig_genes <- colnames(Matrix)
labels_col <- sub("_dupe\\d+$", "", orig_genes)

## 2) Make unique internal column names (adds suffixes like _dupe, _dupe1, …)
cols_unique <- make.unique(orig_genes, sep = "_dupe")
colnames(Matrix) <- cols_unique

## 3) Column annotation: first 31 = Brain, the rest = Heart
n_brain <- min(45, length(orig_genes))
annot_col <- data.frame(
  Organ = factor(c(rep("Brain", n_brain),
                       rep("Heart", length(orig_genes) - n_brain)),
                     levels = c("Brain","Heart"))
)
rownames(annot_col) <- cols_unique  # must match Matrix colnames exactly (and uniquely)

## 4) Row annotation: RNA vs ATAC from the rowname prefix
rows <- rownames(Matrix)
annot_row <- data.frame(
  Method = factor(ifelse(grepl("^RNA\\.", rows), "RNA", "ATAC"),
                  levels = c("RNA","ATAC"))
)
rownames(annot_row) <- rows

## 5) Colors (same as before)
ann_colors <- list(
  Organ = c(Brain = "#3A86FF", Heart = "#2FBF71"),
  Method    = c(RNA   = "red", ATAC  = "purple3")
)
pal_pm1 <- colorRampPalette(c("grey20","grey50","gold3"))(201)

## 6) Optional visual gaps
gap_col <- if (ncol(Matrix) > n_brain) n_brain else NULL
last_rna_idx <- max(which(grepl("^RNA\\.", rows)))
gap_row <- if (length(last_rna_idx)) last_rna_idx else NULL

## 7) Plot, but **show original (duplicated) labels** on the x-axis

# keep original, unique rownames for indexing/annotations
rn <- rownames(Matrix)

# what you want to show on the heatmap axis
labels_row <- sub("^(RNA|ACTIVITY)\\.", "", rn)

pdf("../Results/Manuscript/Figure1/Main1.pdf",width = 17,height = 8)
pheatmap(
  Matrix,
  color              = pal_pm1,
  breaks             = seq(-1, 1, length.out = 201),
  annotation_col     = annot_col,
  annotation_row     = annot_row,
  annotation_colors  = ann_colors,
  cluster_cols       = FALSE,
  cluster_rows       = FALSE,
  gaps_col           = gap_col,
  gaps_row           = gap_row,
  labels_col         = labels_col,   # <- display original symbols
  border_color       = NA,
  main               = "Brain/Heart gene signatures × RNA/ATAC",
  labels_row = labels_row,scale = "none"   # <- shown on the y-axis; rownames stay untouched
)
dev.off()


pdf("../Results/Manuscript/Figure1/Main1(ISS).pdf",width = 12,height = 8)

pheatmap(
  Matrix[,colnames(Matrix)%in%ISS$`Gene Symbol`],
  color              = pal_pm1,
  breaks             = seq(-1, 1, length.out = 201),
  # annotation_col     = annot_col[c(na.omit(match( ISS$`Gene Symbol`,colnames(Matrix))))],
  annotation_row     = annot_row,
  annotation_colors  = ann_colors,
  cluster_cols       = FALSE,
  cluster_rows       = FALSE,
  gaps_col           = 16,
  gaps_row           = gap_row,
  # labels_col         = labels_col,   # <- display original symbols
  border_color       = NA,
  main               = "Brain/Heart gene signatures × RNA/ATAC (ISS genes)",
  labels_row = labels_row,scale = "none"   # <- shown on the y-axis; rownames stay untouched
)
dev.off()

```



Now we want to plot the heart markers in brain and brain markers in heart to select the more organ specific genes

```{r heatmap.fig1,fig.width=25,fig.height=12}
fill_map <- readRDS("../Rsave/fill_map.Rds")


BRAIN <- c("DPEP1","ABCB1","DEPP1","EGFL8","IFI27","GJA4","PLVAP","ESM1","NLGN4Y","HPGD", "TFPI2", "CLEC2B", "CETP","HIST1H3B","SYP","MMP10","MMP1","ADGRG6","PAX3","TOX3","TSLP","RFX4","COL11A1","OGN","CRABP1","CCBE1","HIGD1B","FHL5","SULT1E1","CRYAB","TAGLN","ACTA2","ACTG2","TAGLN","ACTA2","COL8A1","CDC20","MSX2")

HEART <- c("KLHL6","CA4","CLIC5","TSPAN2","GPIHBP1","OIT3","SLC16A13","TNFSF9","ADM","IER5","SMOC1","MYOZ2","ACKR1","MEDAG","TFF3","FLT4","TBX1","LYVE1","LAPTM5","PDPN","MRC1","STON2","STAB2","HIST1H1D","HIST1H2AJ","TYMS","TNFAIP6","PRRX1","CYSLTR2","COL5A3","FOXF1","EGR4")



avg.brain.rna_brain <- as.data.frame(AverageExpression(RNA.all.int$Brain,features = unique(c(BRAIN)),assays = "RNA",group.by = "newL3_Cell_type"))
avg.brain.rna_heart <- as.data.frame(AverageExpression(RNA.all.int$Heart,features = unique(c(BRAIN)),assays = "RNA",group.by = "newL3_Cell_type"))

avg.heart.rna_heart <- as.data.frame(AverageExpression(RNA.all.int$Heart,features = unique(c(HEART)),assays = "RNA",group.by = "newL3_Cell_type"))
avg.heart.rna_brain <- as.data.frame(AverageExpression(RNA.all.int$Brain,features = unique(c(HEART)),assays = "RNA",group.by = "newL3_Cell_type"))

colnames(avg.brain.rna_brain) <- gsub("RNA.","M1.Brain_",colnames(avg.brain.rna_brain))
colnames(avg.brain.rna_heart) <- gsub("RNA.","M1.Heart_",colnames(avg.brain.rna_heart))

colnames(avg.heart.rna_heart) <- gsub("RNA.","M2.Heart_",colnames(avg.heart.rna_heart))
colnames(avg.heart.rna_brain) <- gsub("RNA.","M2.Brain_",colnames(avg.heart.rna_brain))


## --- 1) Row-wise min–max scaler to [-1, 1] (per matrix) ---
scale_rows_to_pm1 <- function(M) {
  # M: genes x samples
  scale_row <- function(x) {
    rmin <- min(x, na.rm = TRUE); rmax <- max(x, na.rm = TRUE)
    if (!is.finite(rmin) || !is.finite(rmax) || rmax == rmin) {
      return(rep(0, length(x)))             # flat row -> all zeros
    }
    2 * ((x - rmin) / (rmax - rmin)) - 1    # map to [-1, 1]
  }
  t(apply(M, 1, scale_row))
}


matrix1 <- as.data.frame(scale_rows_to_pm1(cbind(avg.brain.rna_brain,avg.brain.rna_heart)))
matrix2 <- as.data.frame(scale_rows_to_pm1(cbind(avg.heart.rna_heart,avg.heart.rna_brain)))


## Use your provided palettes
organ_colors <- c(Brain = "#3A86FF", Heart = "#2FBF71")
# fill_map must already be a **named** vector like you showed
# names(fill_map) should match your clean celltype strings exactly

## Helper to normalize celltype spelling once
tidy_ct <- function(x){
  x <- gsub("\\.", "_", x)
  x <- gsub("Tip_cells", "Tip cells", x)
  x <- gsub("Endocardial_like", "Endocardial-like", x)
  x
}

## Use the order given by fill_map (so the legend/order match)
ct_levels <- names(fill_map)

hm_cols   <- colorRampPalette(c("grey20","grey50","white","red2","red4"))(201)
hm_breaks <- seq(-1, 1, length.out = 201)

make_one_heatmap <- function(mtx, matrix_tag = c("M1","M2")) {
  matrix_tag <- match.arg(matrix_tag)
  mtx <- as.matrix(mtx)

  cn <- colnames(mtx)
  # Parse tags: "M1.Brain_ECs.Cpillary" etc.
  organ   <- sub("^[^.]+\\.([^_]+)_.*$", "\\1", cn)
  ctype   <- sub("^[^.]+\\.[^_]+_(.*)$", "\\1", cn) |> tidy_ct()

  ann <- data.frame(
    Organ    = factor(organ, levels = c("Brain","Heart")),
    Celltype = factor(ctype, levels = ct_levels),
    row.names = cn,
    check.names = FALSE
  )

  # Order columns: Brain then Heart within each celltype
  ord <- ann |>
    mutate(Organ_ord = as.integer(factor(Organ, levels = c("Brain","Heart")))) |>
    arrange(Celltype, Organ_ord) |>
    rownames()

  mtx <- mtx[, ord, drop = FALSE]
  ann <- ann[ord, , drop = FALSE]

  # Gaps between each celltype block (2 columns per block)
  run_ct   <- rle(as.character(ann$Celltype))
  gaps_col <- cumsum(run_ct$lengths)
  gaps_col <- gaps_col[-length(gaps_col)]

  ann_colors <- list(
    Organ    = organ_colors,
    Celltype = fill_map
  )

  # Minimal column labels: show Organ under each column (or set show_colnames=FALSE)
  labels_col <- as.character(ann$Organ)

  pheatmap::pheatmap(
    mtx,
    color             = hm_cols,
    breaks            = hm_breaks,
    na_col            = "grey90",
    annotation_col    = ann[, c("Organ","Celltype"), drop = FALSE],
    annotation_colors = ann_colors,
    cluster_cols      = FALSE,
    cluster_rows      = FALSE,
    gaps_col          = gaps_col,
    labels_col        = labels_col,
    fontsize_col      = 8,
    show_rownames     = TRUE,
    main              = if (matrix_tag == "M1")
                          "Brain gene signatures"
                        else
                          "Heart gene signatures"
  )
}

## Render the two separate heatmaps
ht_M1 <- make_one_heatmap(matrix1, matrix_tag = "M1")  # Brain genes
ht_M2 <- make_one_heatmap(matrix2, matrix_tag = "M2")  # Heart genes

pdf("../Results/Manuscript/Figure1/M1_brainGenes_heatmap.pdf", 10, 8); ht_M1; dev.off()
pdf("../Results/Manuscript/Figure1/M2_brainGenes_heatmap.pdf", 10, 8); ht_M2; dev.off()



```



Here I want to see if the Brain specific gene signature are available in heart or not and vice versa

```{r heatmap.fig1,fig.width=25,fig.height=12}

ATAC.all.int <- readRDS(file="./ATAC.all.int2.Rds")
RNA.all.int <- readRDS(file="./RNA.int.Rds")

DefaultAssay(ATAC.all.int) <- "ACTIVITY"
DefaultAssay(RNA.all.int) <- "RNA"

Idents(ATAC.all.int) <- "newL3_Cell_type.multi.pred"
Idents(RNA.all.int) <- "newL3_Cell_type"


BRAIN <- c("DPEP1","ABCB1","DEPP1","EGFL8","IFI27","GJA4","PLVAP","MMP1","MMP10","ESM1","NLGN4Y","HPGD", "TFPI2", "CLEC2B", "CETP", "TMEM233","HIST1H3B","MSX2","FZD2","COL11A1","NUSAP1","CDC20","PAX3","TOX3","TSLP","RFX4","NUPR1","OGN","CRABP1","CCBE1",
"SYP", #Endocardial-like?    
"ACKR1","IL33","ADGRG6","HIGD1B","FHL5","SULT1E1","CRYAB","TAGLN","WFDC1","ACAN","MYL9","ACTA2","ACTG2","TAGLN","MYOCD","ACTA2","COL8A1")

HEART <- c("KLHL6","CA4","CLIC5","EGFL8","TSPAN2","GPIHBP1","PRND","ACKR1","MEDAG","CD69","MADCAM1","OIT3","LXN","SLC16A13","TNFSF9","ADM","IER5","TFF3","FLT4","TBX1","LYVE1","LAPTM5","PDPN","MRC1","STON2","STAB2","HIST1H1D","HIST1H2AL","HIST1H2AJ","TYMS","OMD","TNFAIP6","LUM","PRRX1","SMOC1","MYOZ2","CYSLTR2","AVPR1A","CCDC102B","KCNJ8","COL5A3","FOXF1","PTTG1","EGR4","PSRC1")


DotPlot(ATAC.all.int,features = unique(c(BRAIN,HEART) ),split.by = "Region",cols = c("Brain"="red4","Heart"="red4"))+
  theme_bw() +
  theme(axis.text = element_text(face = "bold", size = 20),
        title     = element_text(face = "bold", size = 20),
        legend.position = "none") +         # <- hide legend for p1
  labs(title = "ATAC-Brain and Heart") + xlab("") + ylab("") +
  rotate_x_text(90)

pdf("../Results/Manuscript/Figure1/Brain_and_heart.vasc.diff.pdf",width = 25,height = 10)

DotPlot(RNA.all.int,features = unique(c(BRAIN,HEART) ),split.by = "Organ",cols = c("Brain"="red4","Heart"="red4"))+
  theme_bw() +
  theme(axis.text = element_text(face = "bold", size = 20),
        title     = element_text(face = "bold", size = 20),
        legend.position = "none") +         # <- hide legend for p1
  labs(title = "ATAC-Brain and Heart") + xlab("") + ylab("") +
  rotate_x_text(90)
dev.off()

cECs <- c("DPEP1","ABCB1","KLHL6","CA4","CLIC5")
Tip <- c("MMP1","MMP10","SLC16A13","TNFSF9")

library(ggplot2)
library(dplyr)
library(tidyr)

genes_use <- Tip
Cell_type="Tip cells" #Tip cells ECs_Cpillary

df <- FetchData(subset(RNA.all.int,newL3_Cell_type==Cell_type), vars = c(genes_use, "newL3_Cell_type", "Organ")) %>%
  pivot_longer(cols = all_of(genes_use), names_to = "gene", values_to = "expr")

df$gene <- factor(df$gene, levels = genes_use)


p2 <- ggplot(df, aes(x = newL3_Cell_type, y = expr)) +
  geom_half_violin(
    data = subset(df, Organ=="Brain"),
    aes(fill = Organ),
    side = "l",width = 1,
    alpha = .85,
    colour = NA
  ) +
  geom_half_violin(
    data = subset(df, Organ=="Heart"),
    aes(fill = Organ),
    side = "r",width = 1,
    alpha = .85,
    colour = NA
  ) +
  scale_fill_manual(values = c(Brain = "#3A86FF", Heart = "#2FBF71")) +
  facet_wrap(~gene, scales="fixed", nrow=1, strip.position="bottom") +
  coord_cartesian(ylim=c(0,4)) +
  theme_bw() +
  theme(
    legend.position = "none",
    strip.background = element_blank(),       # remove facet box
    strip.text = element_text(face="bold", size=16), # gene names as x title
    axis.text.x = element_blank(),            # remove celltype labels
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size=14, face="bold"),
    panel.spacing = unit(0.25, "lines")       # reduces gaps between facets
  ) +
  labs(y="Expression TipC")


pdf("../Results/Manuscript/Figure1/Brain_and_heart.vasc.diff.vln.pdf",width = 8,height = 5)

p1/p2
dev.off()
```




# Motif analysis

Adding the motifs to datasets (has been done in the local computer as packages failed to be installed in the server)

```{r motifs,fig.width=20,fig.height=12}
library(BSgenome.Hsapiens.UCSC.hg38)
library(JASPAR2020)
library(TFBSTools)
library(motifmatchr)

ATAC.all <- readRDS("../Rsave/ATAC.all.int.Splitted.Rds")
DefaultAssay(ATAC.all$Brain) <- "Peak.celltype"
DefaultAssay(ATAC.all$Heart) <- "Peak.celltype"

# Get a list of motif position frequency matrices from the JASPAR database
pfm <- getMatrixSet(
  x = JASPAR2020,
  opts = list(collection = "CORE", tax_group = 'vertebrates',species="Homo sapiens", all_versions = FALSE)
)

# add motif information
ATAC.all$Brain <- AddMotifs(
  object = ATAC.all$Brain,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  pfm = pfm
)

ATAC.all$Heart <- AddMotifs(
  object = ATAC.all$Heart,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  pfm = pfm
)

saveRDS(ATAC.all,"../Rsave//ATAC.all.int.Splitted.Rds")

```


# Differentially enriched fragments

```{r motifs.deg,fig.width=20,fig.height=12}

ATAC.all <- readRDS(file = "../Rsave/ATAC.all.int.Splitted.Rds")
Idents(ATAC.all$Brain) <- "newL3_Cell_type.multi.pred"
Idents(ATAC.all$Heart) <- "newL3_Cell_type.multi.pred"

da_peaks.brain <- FindAllMarkers(
  object = ATAC.all$Brain,
  only.pos = TRUE,
  test.use = 'LR',
  min.pct = 0.05,
  latent.vars = c('nCount_peaks',"Source")
)
saveRDS(da_peaks.brain,"../Rsave/degs.RNA.ATAC.organ.L3_LR_ATACassay.brain.Rds")


# Make covariates clean
ATAC.all$Heart$Source        <- droplevels(factor(ATAC.all$Heart$Source))
ATAC.all$Heart$Source_num    <- as.numeric(ATAC.all$Heart$Source) - 1  # 0/1 batch covariate


da_peaks.heart <- FindAllMarkers(
  object = ATAC.all$Heart,
  only.pos = TRUE,
  test.use = 'LR',
  min.pct = 0.05,
  latent.vars = c('nCount_peaks',"Source_num")
)

saveRDS(da_peaks.heart,"../Rsave/degs.RNA.ATAC.organ.L3_LR_ATACassay.heart.Rds")

```


## Filtering the differentially enriched fragments

Plotting the top enriched motifs per celltype based on the differentially enriched fragments for **brain**

```{r motifs.enrichment.brain,fig.width=20,fig.height=12}

da_peaks.brain <- readRDS("../Rsave/degs.RNA.ATAC.organ.L3_LR_ATACassay.brain.Rds")
dfs <- split(da_peaks.brain,f = da_peaks.brain$cluster,drop = T)

tops <- list()
for (clust in unique(da_peaks.brain$cluster)) {
  set <- subset(da_peaks.brain,cluster==clust)
  tops[[clust]] <- rownames(set[set$p_val < 0.005 & set$pct.1 > 0.2, ])
}

enr.motif.list <- list()

for (name in names(tops)) {
  # test enrichment
enr.motif.list[[name]] <- FindMotifs(
  object = ATAC.all$Brain,assay = "Peak.celltype",
  features = tops[[name]])
}


motif.name.list.brain <- list()
motif.list.brain <- c()
plot.list <- list()
for (name in names(enr.motif.list)) {
plot.list[[name]] <-  MotifPlot(
  object = ATAC.all$Brain,
  motifs = head(rownames(enr.motif.list[[name]])),
)+ggtitle(name)
  motif.list.brain <- c(motif.list.brain, head(enr.motif.list[[name]]$motif.name))
  motif.name.list.brain[[name]] <- head(enr.motif.list[[name]])
}

pdf("../Results/Manuscript/Figure1/Motif.enrcih.brain.pdf",width = 20,height = 15)
wrap_plots(plot.list,ncol =2 )
dev.off()

```


Plotting the top enriched motifs per celltype based on the differentially enriched fragments for **heart**


```{r motifs.enrichment.heart,fig.width=20,fig.height=12}

da_peaks.heart <- readRDS("../Rsave/degs.RNA.ATAC.organ.L3_LR_ATACassay.heart.Rds")
dfs <- split(da_peaks.heart,f = da_peaks.heart$cluster,drop = T)

tops <- list()
for (clust in unique(da_peaks.heart$cluster)) {
  set <- subset(da_peaks.heart,cluster==clust)
  tops[[clust]] <- rownames(set[set$p_val < 0.005 & set$pct.1 > 0.1, ])
}

peak_ids <- rownames(ATAC.all$Heart)

sanitize_peaks <- function(v) {
  v <- as.character(v)
  v <- gsub("\\s+", "", v)
  # strip any trailing digits appended after the end coordinate
  v <- sub("^((chr[^-]+)-(\\d+)-(\\d+))\\d+$", "\\1", v)
  # keep strict "chr-START-END" form
  keep <- grepl("^chr[^-]+-\\d+-\\d+$", v)
  v[keep]
}

keep_std <- function(v) v[grepl("^chr([0-9]{1,2}|X|Y)$", sub("-.*","", v))]

enr.motif.list.heart <- list()
for (name in names(tops)) {
  feats <- sanitize_peaks(tops[[name]])
  feats <- keep_std(feats)
  feats <- intersect(feats, peak_ids)

  if (length(feats) < 50L) {
    message("Skipping ", name, " (", length(feats), " valid peaks)")
    next
  }

  enr.motif.list.heart[[name]] <- tryCatch(
    FindMotifs(object = ATAC.all$Heart, features = feats),
    error = function(e) { message("FindMotifs failed for ", name, ": ", e$message); NULL }
  )
}


motif.name.list.heart <-list() 
motif.list.heart <- c()
plot.list <- list()
for (name in names(enr.motif.list.heart)) {
plot.list[[name]] <-  MotifPlot(
  object = ATAC.all$Heart,
  motifs = head(rownames(enr.motif.list.heart[[name]])),
)+ggtitle(name)

  motif.list.heart <- c(motif.list.heart, head(enr.motif.list.heart[[name]]$motif.name))
  motif.name.list.heart[[name]] <- head(enr.motif.list.heart[[name]])
}

pdf("../Results/Manuscript/Figure1/Motif.enrcih.heart.pdf",width = 20,height = 15)
wrap_plots(plot.list,ncol =2 )
dev.off()

save(motif.name.list.brain,motif.name.list.heart,enr.motif.list.heart,enr.motif.list,file = "../Rsave/motif.top.Rdata")

```


# Runing the motif activity assay=chromvar (degs for motifs) + Checking the shared brain|heart motif enrichment 

```{r motif.act,fig.width=20,fig.height=12}
load("../Rsave/motif.top.Rdata")

ATAC.all$Brain <- RunChromVAR(
  object = ATAC.all$Brain,
  genome = BSgenome.Hsapiens.UCSC.hg38
)

ATAC.all$Heart <- RunChromVAR(
  object = ATAC.all$Heart,
  genome = BSgenome.Hsapiens.UCSC.hg38
)

DefaultAssay(ATAC.all$Brain) <- 'chromvar'
DefaultAssay(ATAC.all$Heart) <- 'chromvar'

################################################################
################################################################

plot.list1 <- list()
plot.list2 <- list()

for (name in intersect(names(enr.motif.list.heart),names(enr.motif.list))) {
 gene <- intersect(enr.motif.list.heart[[name]]$motif[1:20], enr.motif.list[[name]]$motif[1:20])
 gene <- gene[1:3]
p1 <-   MotifPlot(
  object = ATAC.all$Brain,assay = "Peak.celltype",
  motifs = gene,
)+ggtitle(paste0("Brain_",name))

p2 <- MotifPlot(
  object = ATAC.all$Heart,assay = "Peak.celltype",
  motifs = gene,
)+ggtitle(paste0("Heart_",name))

p3 <- FeaturePlot(
  object = ATAC.all$Brain,
  features = gene,
  min.cutoff = 'q10',
  max.cutoff = 'q90',
  pt.size = 0.1,ncol = 3
)&labs(subtitle = paste0("Brain_",name))

p4 <- FeaturePlot(
  object = ATAC.all$Heart,
  features = gene,
  min.cutoff = 'q10',
  max.cutoff = 'q90',
  pt.size = 0.1,ncol = 3
)&labs(subtitle = paste0("Heart_",name))

plot.list1[[name]] <- p1+p2
plot.list2[[name]] <-  wrap_plots(p3,p4,ncol = 2)
}


pdf("../Results/Manuscript/Figure1/Fig1.Supp1.Motif.enrcih.brain_heart.shared.pdf",width = 15,height = 13)
wrap_plots(plot.list1,ncol = 1)
dev.off()


pdf("../Results/Manuscript/Figure1/DIM.Motif.enrcih.brain_heart.shared.pdf",width = 15,height = 25)
wrap_plots(plot.list2,ncol = 1)
dev.off()


##-------------------------------

differential.activity.brain <- FindAllMarkers(
  object = ATAC.all$Brain,
  only.pos = TRUE,
  mean.fxn = rowMeans,
  fc.name = "avg_diff"
)

differential.activity.heart <- FindAllMarkers(
  object = ATAC.all$Heart,
  only.pos = TRUE,
  mean.fxn = rowMeans,
  fc.name = "avg_diff"
)
save(differential.activity.brain,differential.activity.heart,file = "./Diff_Act.Rdata")



## Testing if this way we can have better differential motif 
differential.activity.brain <- FindAllMarkers(
  object = ATAC.all$Brain,assay = "chromvar",test.use = "wilcox",
  only.pos = TRUE
)

differential.activity.heart <- FindAllMarkers(
  object = ATAC.all$Heart,assay = "chromvar",test.use = "wilcox",
  only.pos = TRUE
)

save(differential.activity.brain,differential.activity.heart,file = "./Diff_Act.wilcox.Rdata")

```



```{r motif.act,fig.width=20,fig.height=15}

load("../Rsave/Diff_Act.wilcox.Rdata")
levels <- c("ECs_Cpillary" , "ECs_Arterial" , "Tip cells",  "Endocardial-like","ECs_Venous"  ,"ECs_Lymphatic" ,   "ECs_Mitotic", "Mural_Precursors" , "Mural_Fibroblasts" , "Mural_Pericytes" , "Mural_SMCs","Mural_Mitotic" )

ATAC.all$Brain$newL3_Cell_type.multi.pred <- factor(ATAC.all$Brain$newL3_Cell_type.multi.pred,levels = levels)
differential.activity.brain$cluster <- factor(differential.activity.brain$cluster,levels = levels(ATAC.all$Brain$newL3_Cell_type.multi.pred))
differential.activity.brain <- differential.activity.brain[order(differential.activity.brain$cluster),] 

ATAC.all$Heart$newL3_Cell_type.multi.pred <- factor(ATAC.all$Heart$newL3_Cell_type.multi.pred,levels = levels)
differential.activity.heart$cluster <- factor(differential.activity.heart$cluster,levels = levels(ATAC.all$Heart$newL3_Cell_type.multi.pred))
differential.activity.heart <- differential.activity.heart[order(differential.activity.heart$cluster),] 


get_motif_map <- function(seu, assay_with_motifs = NULL) {
  if (is.null(assay_with_motifs)) {
    # pick the first ChromatinAssay that actually has motifs
    assay_with_motifs <- names(Filter(
      function(a) inherits(a, "ChromatinAssay") && !is.null(a@motifs),
      seu@assays
    ))[1]
  }
  stopifnot(!is.null(assay_with_motifs))
  mot <- seu@assays[[assay_with_motifs]]@motifs
  stopifnot(!is.null(mot))
  # named character vector: names = motif IDs (e.g. "MA0868.2"), values = TF names (e.g. "SOX8")
  mot@motif.names
}



motif2tf_brain <- get_motif_map(ATAC.all$Brain)  # uses the assay that holds motifs
motif2tf_heart <- get_motif_map(ATAC.all$Heart)

## Add TF names to the chromVAR differential tables
differential.activity.brain <- differential.activity.brain %>%
  mutate(TF = unname(motif2tf_brain[gene]),
         TF = ifelse(is.na(TF) | TF == "", gene, TF))  # fallback to motif ID if no name

differential.activity.heart <- differential.activity.heart %>%
  mutate(TF = unname(motif2tf_heart[gene]),
         TF = ifelse(is.na(TF) | TF == "", gene, TF))


# tops.atac.brain <- c()
# for (clusters in unique(differential.activity.brain$cluster)) {
#   x <- subset(differential.activity.brain, cluster == clusters & p_val_adj<0.05)
#   x <- x[order(-x$avg_diff, -x$p_val_adj),][1:5,]
#   tops.atac.brain <- rbind(tops.atac.brain,x)
# }


tops.atac.brain <- c()
for (clusters in unique(differential.activity.brain$cluster)) {
  x <- subset(differential.activity.brain, cluster == clusters & p_val_adj<0.05)
  x <- x[order(-x$avg_log2FC, -x$p_val_adj),][1:5,]
  tops.atac.brain <- rbind(tops.atac.brain,x)
}

differential.activity.brain$TF <- as.character(differential.activity.brain$TF)
differential.activity.heart$TF <- as.character(differential.activity.heart$TF)

top10_tf_per_cluster_brain <- differential.activity.brain %>%
  filter(p_val_adj < 0.01) %>%
  group_by(cluster,gene, TF) %>%
  slice_max(order_by = avg_log2FC, n = 1, with_ties = FALSE) %>%  # keep best motif per TF
  ungroup() %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 5, with_ties = FALSE) %>%
  arrange(cluster, desc(avg_log2FC))

top10_tf_per_cluster_heart <- differential.activity.heart %>%
  filter(p_val_adj < 0.01) %>%
  group_by(cluster, TF) %>%
  slice_max(order_by = avg_log2FC, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 5, with_ties = FALSE) %>%
  arrange(cluster, desc(avg_log2FC))

"%Nin%" <- Negate("%in%")


avg_z.brain <- as.matrix(AverageExpression(
  ATAC.all$Brain,features = unique(top10_tf_per_cluster_brain$gene),
  assays = "chromvar", slot = "data",
  group.by = "newL3_Cell_type.multi.pred",
  return.seurat = FALSE, verbose = FALSE
)$chromvar)   # motifs × clusters (already Z per cell, now averaged)

avg_z.heart <- as.matrix(AverageExpression(
  ATAC.all$Heart,features = unique(top10_tf_per_cluster_heart$gene),
  assays = "chromvar", slot = "data",
  group.by = "newL3_Cell_type.multi.pred",
  return.seurat = FALSE, verbose = FALSE
)$chromvar)   # motifs × clusters (already Z per cell, now averaged)


spec_z.brain <- t(scale(t(as.matrix(avg_z.brain))))  # center/scale per motif across clusters
spec_z.brain[is.na(spec_z.brain)] <- 0               # constant motifs -> NA; set to 0 or drop

spec_z.heart <- t(scale(t(as.matrix(avg_z.heart))))  # center/scale per motif across clusters
spec_z.heart[is.na(spec_z.heart)] <- 0               # constant motifs -> NA; set to 0 or drop

scaled_long.brain <- as_tibble(spec_z.brain, rownames = "TF") |>
  pivot_longer(-TF, names_to = "Domain", values_to = "zscore")

scaled_long.heart <- as_tibble(spec_z.heart, rownames = "TF") |>
  pivot_longer(-TF, names_to = "Domain", values_to = "zscore")

# 5) Quick QC plot
hist(
  scaled_long.brain$zscore,
  col = "skyblue", border = "white",
  main = "Specificity (Z) Scores — merged domains",
  xlab = "Z-score"
)
abline(v = c( 2), col = "red", lwd = 2)

hist(
  scaled_long.heart$zscore,
  col = "skyblue", border = "white",
  main = "Specificity (Z) Scores — merged domains",
  xlab = "Z-score"
)
abline(v = c( 2), col = "red", lwd = 2)



scaled_long.brain <- subset(scaled_long.brain,zscore>2)
top10_tf_per_cluster_brain <- subset(top10_tf_per_cluster_brain,gene %in% scaled_long.brain$TF)

scaled_long.heart <- subset(scaled_long.heart,zscore>2)
top10_tf_per_cluster_heart <- subset(top10_tf_per_cluster_heart,gene %in% scaled_long.heart$TF)



library(grid)
library(patchwork)

p1 <- DotPlot(
  subset(ATAC.all$Brain, newL3_Cell_type.multi.pred %Nin% c("ECs_Mitotic","ECs_Lymphatic")),
  scale.by = "radius", assay = "chromvar",
  features = unique(top10_tf_per_cluster_brain$gene),
  group.by = "newL3_Cell_type.multi.pred"
) +
  coord_flip() +
  scale_x_discrete(labels = top10_tf_per_cluster_brain$TF) +
  scale_color_gradient2(low = "white", mid = "lightblue", high = "purple3") +
  theme_bw() +
  theme(axis.text = element_text(face = "bold", size = 20),
        title     = element_text(face = "bold", size = 20),
        legend.position = "none") +         # <- hide legend for p1
  labs(title = "ATAC-Brain",subtitle = "Motif activity") + xlab("") + ylab("") +
  rotate_x_text(90)

p2 <- DotPlot(
  subset(ATAC.all$Heart, newL3_Cell_type.multi.pred %Nin% c("ECs_Venous","Mural_SMCs")),
  scale.by = "radius", assay = "chromvar",
  features = unique(top10_tf_per_cluster_heart$gene),
  group.by = "newL3_Cell_type.multi.pred"
) +
  coord_flip() +
  scale_x_discrete(labels = top10_tf_per_cluster_heart$TF) +
  scale_color_gradient2(low = "white", mid = "lightblue", high = "purple3") +
  theme_bw() + xlab("") + ylab("") +
  rotate_x_text(90) +
  labs(title ="ATAC-Heart" ,color = "Activity", size = "% cells",subtitle = "Motif activity") +
  guides(
    color = guide_colorbar(title.position = "top",
                           barheight = unit(5, "cm"),
                           barwidth  = unit(0.35, "cm")),
    size  = guide_legend(title.position = "top", ncol = 1,
                         override.aes = list(alpha = 1))
  ) +
  theme(axis.text = element_text(face = "bold", size = 20),
        title     = element_text(face = "bold", size = 20),
        legend.position = "right",           # <- keep legend for p2 on the right
        legend.direction = "vertical")


pdf("../Results/Manuscript/Figure1/TFenrichment.wilcox.pdf",width = 12,height = 11)
p1+p2
dev.off()




## --- Brain ---

obj_b <- subset(ATAC.all$Brain, newL3_Cell_type.multi.pred %Nin% c("ECs_Mitotic","ECs_Lymphatic"))
grp_levels_b <- levels(obj_b$newL3_Cell_type.multi.pred)

# Main dot plot (no y text; we’ll show a color strip instead)
p1_core <- DotPlot(
  obj_b,
  scale.by = "radius", assay = "chromvar",
  features = unique(top10_tf_per_cluster_brain$gene),
  group.by = "newL3_Cell_type.multi.pred"
) +
  coord_flip() +
  scale_x_discrete(labels = top10_tf_per_cluster_brain$TF) +
  scale_color_gradient2(low = "white", mid = "lightblue", high = "purple3") +
  theme_bw() +
  theme(
    axis.title  = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.y   = element_blank(),
    legend.position= "none",
    plot.margin    = margin(2, 4, 0, 4) # tighter bottom to snug the strip
  ) +
  labs(title = "ATAC - Brain", subtitle = "Motif activity")

# Thin horizontal color strip UNDER the plot (no text)
strip_b_bottom <- ggplot(
  data.frame(grp = factor(grp_levels_b, levels = grp_levels_b)),
  aes(x = grp, y = 1, fill = grp)
) +
  geom_tile(height = 0.25) +                      # make it thin
  scale_fill_manual(values = fill_map, drop = FALSE) +
  scale_y_continuous(expand = c(0,0)) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.margin     = margin(0, 4, 4, 4)          # tiny gap below the dotplot
  )

# Stack dotplot over strip (very short strip)
p1 <- p1_core / strip_b_bottom + plot_layout(heights = c(1, 0.07))


## --- Heart ---

obj_h <- subset(ATAC.all$Heart, newL3_Cell_type.multi.pred %Nin% c("ECs_Venous","Mural_SMCs"))
grp_levels_h <- levels(obj_h$newL3_Cell_type.multi.pred)

p2_core <- DotPlot(
  obj_h,
  scale.by = "radius", assay = "chromvar",
  features = unique(top10_tf_per_cluster_heart$gene),
  group.by = "newL3_Cell_type.multi.pred"
) +
  coord_flip() +
  scale_x_discrete(labels = top10_tf_per_cluster_heart$TF) +
  scale_color_gradient2(low = "white", mid = "lightblue", high = "purple3") +
  theme_bw() +
  theme(
    axis.title      = element_blank(),
      axis.text.x  = element_blank(),
    axis.ticks.y    = element_blank(),
    legend.position = "right"
  ) +
  labs(title = "ATAC - Heart", subtitle = "Motif activity",
       color = "Activity", size = "% cells")

strip_h_bottom <- ggplot(
  data.frame(grp = factor(grp_levels_h, levels = grp_levels_h)),
  aes(x = grp, y = 1, fill = grp)
) +
  geom_tile(height = 0.25) +
  scale_fill_manual(values = fill_map, drop = FALSE) +
  scale_y_continuous(expand = c(0,0)) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.margin     = margin(0, 4, 4, 4)
  )

p2 <- p2_core / strip_h_bottom + plot_layout(heights = c(1, 0.07))

library(patchwork)

final <- (p1 | p2) +
  plot_layout(guides = "collect") 
final

pdf("../Results/Manuscript/Figure1/TFenrichment.wilcox_2.pdf",width = 9,height = 9)
final
dev.off()
```


I want to make module scores for lineage associated genes for heart and brain separately to show that these genes are exclusive to one organ


```{r motif.act,fig.width=20,fig.height=15}

RNA.all.int.all <- readRDS(file="./RNA.int.Rds")
RNA.all.int <- subset(RNA.all.int.all,newL3_Cell_type%in%c("ECs_Arterial",   "ECs_Cpillary", "ECs_Venous"))

RNA.all.int <- SplitObject(RNA.all.int,split.by = "newL3_Cell_type")


aEC_genes <- c("SERPINE2","APOA1","GRIA2","TSPAN2","EGFL8","BTNL9","JAG2","GJA4","PRND","PIK3R3","ARL15","SLC45A4","DLL4","C12orf75","SEMA3G","UNC5B","HEY1")

cEC_genes <- c("CD9","APCDD1","SLC38A5","SLC16A1","HOPX","CHGA","SLC7A8","CDH6","TNNT3","QRFPR","MEOX1","CLIC5","VAV3","HECW2","CD36")

vEC_genes <- c("NCOA7","C11orf96","POSTN","COL3A1","CCL2","IRF1","CEBPD","MMRN1","CCL14","PLVAP","ADGRG6","SELE","ICAM1","FNIP2")


RNA.all.int$ECs_Arterial <- AddModuleScore(RNA.all.int$ECs_Arterial,features = aEC_genes)
RNA.all.int$ECs_Venous <- AddModuleScore(RNA.all.int$ECs_Venous,features = vEC_genes)
RNA.all.int$ECs_Cpillary <- AddModuleScore(RNA.all.int$ECs_Cpillary,features = cEC_genes)

colnames(RNA.all.int$ECs_Arterial@meta.data)[106:122] <- aEC_genes
colnames(RNA.all.int$ECs_Venous@meta.data)[106:119] <- vEC_genes
colnames(RNA.all.int$ECs_Cpillary@meta.data)[106:120] <- cEC_genes



gene.list= list("First"=cEC_genes,
            "Second"=aEC_genes,"Third"=vEC_genes)
library(dplyr)

pics <- list()

for (n in 1:3) {
df <- RNA.all.int[[n]]@meta.data %>%
  dplyr::select(all_of(gene.list[[n]]), Organ) %>%
  tidyr::pivot_longer(cols = all_of(gene.list[[n]]),
                      names_to = "gene",
                      values_to = "score")

df2 <- df %>%
  group_by(gene, Organ) %>%
  summarize(mean = mean(score), .groups = 'drop') %>%
  tidyr::pivot_wider(names_from = Organ, values_from = mean)

df2$gene <- factor(df2$gene,levels=rev(gene.list[[n]]))

p <- ggplot(df2, aes(y = gene)) +
  geom_point(aes(x = Brain, color="Brain"), size=3) +
  geom_point(aes(x = Heart, color="Heart"), size=3) +
  geom_segment(aes(x=Brain, xend=Heart, y=gene, yend=gene), size=0.6, alpha=0.6) +
  scale_color_manual(values=c(Brain="#3A86FF", Heart="#2FBF71")) +
  coord_cartesian(xlim=c(-1,2.1)) +
  theme_classic()+xlab("Gene score")+ggtitle(names(RNA.all.int)[[n]])+ylab("")

pics[[n]] <-p 
}

pics[[1]] <- pics[[1]]+ylab("Average of gene scores")+NoLegend()
pics[[2]] <- pics[[2]]+NoLegend()

pdf("../Results/Manuscript/Figure3/Gene.lineage.score.pdf",width = 12,height = 4)

wrap_plots( pics,ncol = 3)
dev.off()


RNA.all.int <- subset(RNA.all.int.all,newL3_Cell_type%in%c("Mural_Pericytes", "Mural_SMCs"))
RNA.all.int <- SplitObject(RNA.all.int,split.by = "newL3_Cell_type")

PCs_genes <- unique(c("KCNJ8", "TESC", "CCL2","ABCC9", "ADAMTSL3", "ARHGDIB", "COL6A3", "COLEC11", "CYGB", "LAMC3", "SPON2", "THBS4"))
SMCs_genes <- unique(c("ACTA2", "BCL2", "DMD", "ENOX1", "INPP4B", "PARD3B", "PDLIM5", "SLC8A1", "SLIT3", "SOX5", "SYNPO2", "SYTL2", "TAGLN", "TEX41", "UTRN",
                "ACTG2", "ADIRF", "CASQ2", "CNN1", "CRYAB", "ELN", "ERBB4", "LAMA3", "LTBP1", "MYH11", "MYOCD", "RGS6", "SDK1", "SLIT3"))

RNA.all.int$Mural_Pericytes <- AddModuleScore(RNA.all.int$Mural_Pericytes,features = PCs_genes)
RNA.all.int$Mural_SMCs <- AddModuleScore(RNA.all.int$Mural_SMCs,features = SMCs_genes)

colnames(RNA.all.int$Mural_Pericytes@meta.data)[106:117] <- PCs_genes
colnames(RNA.all.int$Mural_SMCs@meta.data)[106:133] <- SMCs_genes


gene.list= list("First"=PCs_genes,
            "Second"=SMCs_genes)
library(dplyr)

pics <- list()

for (n in 1:2) {
df <- RNA.all.int[[n]]@meta.data %>%
  dplyr::select(all_of(gene.list[[n]]), Organ) %>%
  tidyr::pivot_longer(cols = all_of(gene.list[[n]]),
                      names_to = "gene",
                      values_to = "score")

df2 <- df %>%
  group_by(gene, Organ) %>%
  summarize(mean = mean(score), .groups = 'drop') %>%
  tidyr::pivot_wider(names_from = Organ, values_from = mean)

df2$gene <- factor(df2$gene,levels=rev(gene.list[[n]]))

p <- ggplot(df2, aes(y = gene)) +
  geom_point(aes(x = Brain, color="Brain"), size=3) +
  geom_point(aes(x = Heart, color="Heart"), size=3) +
  geom_segment(aes(x=Brain, xend=Heart, y=gene, yend=gene), size=0.6, alpha=0.6) +
  scale_color_manual(values=c(Brain="#3A86FF", Heart="#2FBF71")) +
  # coord_cartesian(xlim=c(-1,2.1)) +
  theme_classic()+xlab("Gene score")+ggtitle(names(RNA.all.int)[[n]])+ylab("")

pics[[n]] <-p 
}

pics[[1]] <- pics[[1]]+ylab("Average of gene scores")+NoLegend()
pics[[2]] <- pics[[2]]+NoLegend()

pdf("../Results/Manuscript/Figure3/Gene.lineage.score.mural.pdf",width = 5,height = 9)

pics[[1]] / pics[[2]] + plot_layout(heights = c(1, 2))
dev.off()

```


For the extended figure 1 I want to calculate the number of cells and samples per organ and timepoint (PCW)

Besides, here are the QC for the extended figures

```{r motif.act,fig.width=20,fig.height=15}


ATAC.all.int <- readRDS(file="./ATAC.all.int2.Rds")
RNA.all.int <- readRDS(file="./RNA.int.Rds")
RNA.all.int <- subset(RNA.all.int, Age.PCW != "PCW22")

ATAC.all.int$Organ <- ATAC.all.int$Region

cells_tbl.RNA <- RNA.all.int@meta.data %>%
  count(Organ, Age.PCW, name = "n_cells")

cells_tbl.ATAC <- ATAC.all.int@meta.data %>%
  count(Organ, Age.PCW, name = "n_cells")

samples_tbl.RNA <- RNA.all.int@meta.data %>%
  distinct(Organ, Age.PCW, orig.ident) %>%
  count(Organ, Age.PCW, name = "n_samples")

samples_tbl.ATAC <- ATAC.all.int@meta.data %>%
  distinct(Organ, Age.PCW, orig.ident) %>%
  count(Organ, Age.PCW, name = "n_samples")

summary_tbl.RNA <- left_join(
  cells_tbl.RNA,
  samples_tbl.RNA,
  by = c("Organ", "Age.PCW")
)

summary_tbl.ATAC <- left_join(
  cells_tbl.ATAC,
  samples_tbl.ATAC,
  by = c("Organ", "Age.PCW")
)


summary_wide.RNA <- summary_tbl.RNA %>%
  pivot_wider(
    names_from = Organ,
    values_from = c(n_cells, n_samples)
  )


summary_wide.ATAC <- summary_tbl.ATAC %>%
  pivot_wider(
    names_from = Organ,
    values_from = c(n_cells, n_samples)
  )




p1 <- ggplot(summary_tbl.RNA, aes(x = Age.PCW, y = n_cells, fill = Organ)) +
  geom_col(position = "stack") +
  geom_point(
    aes(size = n_samples),
    color = "black",
    position = position_stack(vjust = 0.5)
  ) +
  scale_fill_manual(values = c(Brain = "#3A86FF", Heart = "#2FBF71")) +
  scale_size(range = c(2,6)) +
  theme_bw() +
  theme(
    axis.text = element_text(face="bold"),
    axis.title = element_text(face="bold"),
    legend.position = "right"
  ) +
  labs(
    y = "Number of cells",
    x = "Developmental age (PCW)",
    size = "Number of samples",
    title = "RNA level"
  )+rotate_x_text(45)

p2 <- ggplot(summary_tbl.ATAC, aes(x = Age.PCW, y = n_cells, fill = Organ)) +
  geom_col(position = "stack") +
  geom_point(
    aes(size = n_samples),
    color = "black",
    position = position_stack(vjust = 0.5)
  ) +
  scale_fill_manual(values = c(Brain = "#3A86FF", Heart = "#2FBF71")) +
  scale_size(range = c(2,6)) +
  theme_bw() +
  theme(
    axis.text = element_text(face="bold"),
    axis.title = element_text(face="bold"),
    legend.position = "right"
  ) +
  labs(
    y = "Number of cells",
    x = "Developmental age (PCW)",
    size = "Number of samples",
    title = "ATAC level"
  )+rotate_x_text(45)


pdf("../Results/Manuscript/Figure1/Statistics.pdf",width = 10,height =8)
p1+p2
dev.off()


## Number of samples per Organ
RNA <- as.data.frame(table(RNA.all.int$Organ,RNA.all.int$orig.ident))
ATAC <- as.data.frame(table(ATAC.all.int$Organ,ATAC.all.int$orig.ident))
RNA <- RNA[RNA$Freq != 0,]
ATAC <-  ATAC[ATAC$Freq != 0,]

write.csv(summary_wide.RNA,"../Results/Manuscript/Figure1/Statistics.RNA.csv")
write.csv(summary_wide.ATAC,"../Results/Manuscript/Figure1/Statistics.ATAC.csv")


######### QC
DefaultAssay(RNA.all.int) <- "RNA"

p1 <- FeatureScatter(
    RNA.all.int,
    feature1 = "nCount_RNA",
    feature2 = "nFeature_RNA",
    group.by = "Organ",cols = c(Brain = "#3A86FF", Heart = "#2FBF71"),split.by = "Age.PCW",ncol = 5,pt.size = 0.1
) +
    theme_bw() +
    theme(
        axis.text  = element_text(face = "bold", size = 12),
        axis.title = element_text(face = "bold", size = 14)
    ) +
    labs(
        x = "Number of UMIs",
        y = "Number of detected genes",
        title = "QC for RNA level"
    )+rotate_x_text(30)



pdf("../Results/Manuscript/Final_refinments/ED1/QC.RNA.pdf",width = 10,height =5)
p1
dev.off()





DefaultAssay(ATAC.all.int) <- "ATAC"


ATAC.all.int <- TSSEnrichment(ATAC.all.int)
ATAC.all.int <- readRDS("../Rsave/ATAC.all.int2_TSS.Rds")
ATAC.all.int$Source_Region <- gsub("Sten_Brain","MannensC.2023_Brain",ATAC.all.int$Source_Region)
ATAC.all.int$Source_Region <- gsub("AmeenM_Heart","AmeenM.2023_Heart",ATAC.all.int$Source_Region)
ATAC.all.int$Organ <-ATAC.all.int$Region



library(gghalves)

lv <- c("PCW06","PCW07","PCW08","PCW09","PCW10","PCW12","PCW13")
df  <- df %>% mutate(Age.PCW = factor(Age.PCW, levels = lv))

p1 <- ggplot(df, aes(x = Age.PCW, y = TSS.enrichment)) +
  # left half: Heart
  gghalves::geom_half_violin(
    data = ~ subset(.x, Organ == "Heart") |>
            transform(Age.PCW = factor(Age.PCW, levels = lv)),
    aes(fill = "Heart"),
    side = "l", width = 0.9, trim = TRUE, alpha = 0.9
  ) +
  # right half: Brain
  gghalves::geom_half_violin(
    data = ~ subset(.x, Organ == "Brain") |>
            transform(Age.PCW = factor(Age.PCW, levels = lv)),
    aes(fill = "Brain"),
    side = "r", width = 0.9, trim = TRUE, alpha = 0.9
  ) +
  # optional half boxplots
  gghalves::geom_half_boxplot(
    data = ~ subset(.x, Organ == "Heart") |>
            transform(Age.PCW = factor(Age.PCW, levels = lv)),
    aes(fill = "Heart"),
    side = "l", width = 0.5, outlier.shape = NA
  ) +
  gghalves::geom_half_boxplot(
    data = ~ subset(.x, Organ == "Brain") |>
            transform(Age.PCW = factor(Age.PCW, levels = lv)),
    aes(fill = "Brain"),
    side = "r", width = 0.5, outlier.shape = NA
  ) +
  scale_x_discrete(limits = lv, drop = FALSE) +  # lock order
  scale_fill_manual(values = c(Brain = "#3A86FF", Heart = "#2FBF71"),
                    breaks = c("Brain","Heart"), name = "Organ") +
  labs(x = "Age (PCW)", y = "TSS enrichment") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "top")+ylim(0,25)



pdf("../Results/Manuscript/Final_refinments/ED1/ATAC.TSS.enrichmentQC.pdf",width = 8,height = 5)
p1
dev.off()

ATAC.all.int$Organ <-ATAC.all.int$Region


obj <- ATAC.all.int
DefaultAssay(obj) <- "ATAC"

qc_na <- with(obj@meta.data, is.na(nCount_peaks) | is.na(nFeature_peaks) |
                               is.na(FRIP) | is.na(TSS.enrichment) |
                               is.na(nucleosome_signal) | is.na(blacklist_ratio) |
                               is.na(PromoterRatio))
sum(qc_na)
cells_na <- rownames(obj@meta.data)[qc_na]



# CountFragments returns a data.frame with barcode, reads, and duplicates
cf_list <- lapply(Fragments(obj), CountFragments)

cf      <- do.call(rbind, cf_list)
colnames(cf)[1] <- "barcode"
totals  <- setNames(cf$reads, cf$barcode)
obj$total_fragments <- totals[colnames(obj)]
```


